---
title: "Lesson 8"
author: "James B. Elsner"
date: "February 3, 2021"
output:
  html_document: null
editor_options:
  chunk_output_type: console
---

**"Measuring programming progress by lines of code is like measuring aircraft building progress by weight."** – Bill Gates

S4 spatial data objects and working with raster data in R.

## Spatial data frames in the S4 class of data objects

The {sp} package also has methods for working with spatial data. Several of the packages for analyzing and modeling spatial data we will use this semester depend on {sp} so it is important to see how to manipulate these data objects.

Install and load the package.
```{r}
if(!require(sp)) install.packages(pkgs = "sp", repos = "http://cran.us.r-project.org")

library(sp)
```

Spatial objects from the {sp} package fall into two types: 1) spatial-only information (the geometry). These include `SpatialPoints`, `SpatialLines`, `SpatialPolygons`, etc, and 2) extensions to these types where attribute information is available and stored in a data frame. These include `SpatialPointsDataFrame`, `SpatialLinesDataFrame`, etc.

We use `as_Spatial()` to convert a simple feature data frame (as an S3 spatial object) to an S4 spatial object.
```{r}
library(sf)
library(tidyverse)

if(!"1950-2018-torn-aspath" %in% list.files()) {
  download.file(url = "http://myweb.fsu.edu/jelsner/temp/data/1950-2018-torn-aspath.zip",
                destfile = "1950-2018-torn-aspath.zip")
  unzip("1950-2018-torn-aspath.zip")
  }


FL_Torn.sf <- st_read(dsn = "1950-2018-torn-aspath", 
                   layer = "1950-2018-torn-aspath") %>%
  filter(st == "FL")

FL_Torn.sp <- FL_Torn.sf %>%
  as_Spatial()

class(FL_Torn.sp)
```

The result is a S4 spatial object named `FL_Torn.sp` of class `SpatialLinesDataFrame`. 

Data in S4 spatial objects is accessed through a slot name. The slot names are listed with the `slotNames()` function.
```{r}
FL_Torn.sp %>%
  slotNames()
```

The `data` slot contains the data frame, the `lines` slot contains the spatial geometries (in this case lines), the `bbox` slot is the boundary box and the `proj4string` slot is the CRS.

The object name followed by the `@` symbol allows access to the information in the slot. For example to see the first row of the data frame, and the corresponding first spatial geometry type
```{r}
FL_Torn.sp@data[1, ]

FL_Torn.sp@lines[1]
```

The `@` symbol is similar to the `$` symbol for regular data frames.

When using the `$` symbol on S4 spatial objects, we can access the data as a regular data frame. For example, to list the EF rating of all the tornadoes type
```{r}
FL_Torn.sp$mag
```

Selecting, retrieving, or replacing attributes in S4 spatial data frames is done with methods in {base} R package. For example `[]` is used to select rows and/or columns. To select `mag` of the 7th tornado type
```{r}
FL_Torn.sp$mag[7]
```

Other methods include: `plot`, `summary`,`dim` and `names` (operate on the data slot), `as.data.frame`, `as.matrix` and `image` (for spatial data on a grid), and `length` (number of features).

CAUTION: we can't use the {dplyr} verbs on S4 data frames. To convert from an S4 spatial data frame to a simple feature data frame, use `st_as_sf()`.

The interface to the geometry engine-open source (GEOS) is through the {rgeos} package.
```{r}
library(rgeos)
```

When possible we will do our geo-computations on simple feature data frames. However, sometimes it is more convenient to perform geo-computations on S4 data frames. Also legacy R code you find will be written with S4 objects.

To see if the S4 spatial data frame is projected type
```{r}
is.projected(FL_Torn.sp)
```

To see the coordinate reference system (CRS) of the spatial data frame type
```{r}
FL_Torn.sp@proj4string
```

The CRS arguments include the projection (`+proj`), the datum (`+datum`), and the ellipsoid (`+ellps`). Here we see the projection is `longlat` indicating that this is a geographic CRS (not projected).

To create a projected `SpatialLinesDataFrame` we use the `spTransform()` function. The first argument is the original spatial data frame and the second argument is the coordinate reference system as a character string.
```{r}
FL_TornP.sp <- FL_Torn.sp %>%
  spTransform(CRS = CRS("+proj=merc +ellps=GRS80 +units=m"))
is.projected(FL_TornP.sp)
```

The CRS character string is in the open GIS standard format. It includes the projection type (here Mercator), the ellipsoid shape (here GRS80) and the spatial units (here meters).
```{r}
FL_TornP.sp@proj4string
```

We now have two copies of our `SpatialLinesDataFrame` object (unprojected and projected).

We perform geo-computation on the projected spatial data frame using functions from the {rgeos} package. 

Computations can be done across all features (e.g., all tornado reports together) or feature by feature (`byid = TRUE`). For example, to the `gEnvelope()` function computes the rectangular bounding box surrounding all the features.
```{r}
box <- gEnvelope(FL_TornP.sp)
class(box)
```

The assigned object `box` is of class `SpatialPolygons`. It contains a single polygon rectangle. The `byid = FALSE` is the default. There are no attributes.

Plot the box and the tornadoes using {base} R.
```{r}
plot(box)
plot(sldfP, add = TRUE)
```

Note that the `plot()` method applied to an S4 spatial object plots the geometries without the attributes.

As mentioned previously, we will do most of our pre and post-processing using simple feature data frames, so we do not need to dig deeper into the structure of these S4 data objects.

## Working with rasters

The _raster data model_ divides geographic space into a grid of cells of constant size (resolution) and we use classes from the {raster} package to work with raster data.

A raster is a data structure that divides space into rectangles called 'cells' (or 'pixels'). Each cell has an attribute value.

The {raster} package has functions for creating, reading, manipulating, and writing raster data as S4 objects and functions for raster manipulation common in GIS.
```{r}
library(raster)
```

An S4 `RasterLayer` object (raster) is one variable (called a 'layer' or a 'band'). The object includes the number of columns and rows, the coordinates of its spatial extent (bounding box), and the coordinate reference system (CRS). 

A raster can also store information about the external file from where it came.

The package has two classes for data with more than one variable associated with each cell (multi-layer data) the `RasterStack` and the `RasterBrick`. A `RasterBrick` can only be linked to a single (multi-layer) file. A `RasterStack` can be formed from separate files. Otherwise they are the same.

Creating a raster

The `raster()` function creates a raster with a geographic (longitude/latitude) CRS and a 1 by 1 degree grid of cells across the globe.
```{r}
r <- raster()
r
```

Arguments including `xmin`, `nrow`, `ncol`, and `crs` are used to change these default settings. The CRS is written as a proj4string.

The default raster layer is in geographic coordinates spanning the globe at one-degree resolution in the north-south and the east-west directions.

We know the raster object is an S4 class because it has slots (e.g., `@data`).
```{r}
str(r)
```

The slot names are retrieved with
```{r}
slotNames(r)
```

To create an alternative raster with 36 longitudes -100 and 0 degrees East longitude and 18 latitudes between the equator and 50 degrees N latitude we specify the number of columns, the number of rows and the extent as follows.
```{r}
r <- raster(ncol = 36, nrow = 18, 
            xmn = -100, xmx = 0, 
            ymn = 0, ymx = 50)
r
res(r)
```

This results in raster with cell resolution of 2.7 degrees of longitude and 2.7 degrees of latitude.

The parameters can be changed after the raster is created. Here we change the resolution to exactly 3 degrees. This changes the number of rows and columns.
```{r}
res(r) <- 3
ncol(r)
nrow(r)
```

The raster object `r` is a template with no values in the cells. By default it will have an extent that spans the globe.
```{r}
r <- raster(ncol = 10, nrow = 10)
ncell(r)
hasValues(r)
```

Here there are 100 cells in a 10 by 10 empty raster. We use the `values()` function to place values in the cells. The function is specified on the left-hand side of the assignment operator. First we assign to a vector of length `ncell(r)` random numbers from a uniform distribution with the `runif()` function. The default is that the random numbers are between 0 and 1.
```{r}
v <- runif(ncell(r))
head(v)
values(r) <- v
head(r)
```

The cells are arranged in lexicographical order (upper left to lower right) and the cells are populated with the values in the vector in this order.

The `plot()` method creates a choropleth map of the values in cells.
```{r}
plot(r)
```

The default CRS is geographic.
```{r}
projection(r)
```

To project the raster use the function `projectRaster()`. Projections are performed using the `PROJ4` protocol accessed through the {rgdal} package (same as with {sf} and {sp} objects).

We can also use the `setValues()` function to place values in the cells. Here we create a new raster layer with cell numbers as values using the `setValues()` function to place the numbers in the cells.
```{r}
r <- raster(xmn = -110, xmx = -90, 
            ymn = 40, ymx = 60, 
            ncols = 10, nrows = 10)
r <- setValues(r, 1:ncell(r))
projection(r)
plot(r)
```

The numbers increase from top right to bottom left.

Next we create a projected raster.
```{r}
rp <- projectRaster(r, 
                    crs = "+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m")
plot(rp)
```

The projection is performed using bi-linear interpolation (default). In the case where the values are categorical we use `method = "ngb"` for nearest neighbor.

The `raster()` function imports data with functions from the {rgdal} package. Supported formats include `GeoTIFF`, `ESRI`, `ENVI`, and `ERDAS`. Most formats that can import rasters can also be used to exporting rasters. 

Consider the `Meuse` dataset (taken from the {sp} package), using a file in the native 'raster- file' format:
```{r}
f <- system.file("external/test.grd", 
                 package = "raster")
r <- raster(f)
filename(r)
```

Do the cells contain values? Is the raster stored in memory? Create a plot.
```{r}
hasValues(r)
inMemory(r)
plot(r, main = "Raster layer from file")
```

Note the raster layer is a rectangle of cells with values. Values that are coded as `NA` are not plotted.
```{r}
head(r)
```

We can combine raster layers into stacks and bricks. A stack (and a brick) is a collection of raster layers having the same spatial extent and resolution. 

A brick is typically imported from a multi-layer (band) external file. They can also exist entirely in memory. A raster brick can only point to a single external file.

For example, create three rasters and assign random values to the cells.
```{r}
r1 <- r2 <- r3 <- raster(nrow = 10, ncol = 10)
values(r1) <- runif(ncell(r1))
values(r2) <- runif(ncell(r2))
values(r3) <- runif(ncell(r3))
```

Then combine the rasters into a stack with `stack()` (or into a brick with `brick()`).
```{r}
s <- stack(r1, r2, r3)
s
nlayers(s)
plot(s)
```

So each attribute in a raster stack (or brick) is a layer.

Here we import a raster brick from a file.
```{r}
f <- system.file("external/rlogo.grd", 
                 package = "raster")
b <- brick(f)
b
plot(b)
```

Suppose we only want a single layer (e.g., one attribute). We extract a raster layer from a brick or stack with the `layer =` or `band =` arguments.
```{r}
r <- raster(b, layer = 2)
plot(r)
```

Most base R functions (`+`, `*`, `round()`, `ceiling()`, `log()`, etc) work on raster objects. Operations are done on all cells at once.

Here we place the numbers from 1 to 100 sequentially in the cells, then add 100 to these values and take the square root.
```{r}
r <- raster(ncol = 10, nrow = 10)
values(r) <- 1:ncell(r)
s <- r + 100
s <- sqrt(s)
plot(s)
```

Here we replace the cell values with random uniform numbers between 0 and 1. Then round to the nearest integer and add one.
```{r}
r <- setValues(r, runif(ncell(r)))
r <- round(r)
r <- r + 1
plot(r)
```

Replace only certain values with the subset function `[]`.
```{r}
r <- raster(xmn = -90, xmx = 90, ymn = -30, ymx = 30)
values(r) <- rnorm(ncell(r))
plot(r)
s <- raster(xmn = -60, xmx = 60, ymn = -10, ymx = 10)
r[s] <- 0
plot(r)
r[r > 2] <- 0
plot(r)
```

Raster objects with the same resolution and origin can be combined.

To get summary statistics across all values _within_ a particular raster we use the `cellStats()` function.
```{r}
cellStats(r1, stat = 'mean')
cellStats(b, stat = 'sum')
```

## Geo-computation on rasters

The `crop()` function takes a geographic subset of a larger raster object. A raster is cropped by providing an extent object or other spatial object from which an extent can be extracted (objects from classes deriving from raster and from spatial in the {sp} package). 

The function `drawExtent()` is used to visually see the new extent (bounding box) that is given to the `crop()` function.

The `trim()` function crops a raster layer by removing the outer rows and columns that only contain `NA` values. The `extend()` function adds new rows and/or columns with `NA` values.

The `extract()` function retrieves values from a raster object at the locations of other spatial data.

The `merge()` function combines two or more objects into a single object. The input objects must have the same resolution and origin (such that their cells fit into a single larger raster). If this is not the case, first adjust one of the objects with the functions `aggregate()` or `resample()`.

The `aggregate()` and `disaggregate()` functions change the resolution (cell size) of a raster object.
```{r}
r <- raster()
r[] <- 1:ncell(r)
ra <- aggregate(r, 10)
```

Here we crop the raster into two pieces and then merge them into one with the `merge()` function. The function has an argument that allows us to export to `test.grd`.
```{r}
r1 <- crop(r, extent(-180, 0, 0, 30)) 
r2 <- crop(r, extent(-10, 180, -20, 10))
m <- merge(r1, r2, 
           filename = 'test.grd', 
           overwrite = TRUE)
plot(m)
```

Other formats for saving the raster (e.g., `geoTIFF`) are available (see `writeRaster()` for more options). 

The `flip()` function flips the data (reverse order) in the horizontal or vertical direction---typically to correct for a 'communication problem' between different R packages or a misinterpreted file. The `rotate()` function rotates longitude/latitude rasters that have longitudes from 0 to 360 degrees (often used by climatologists) to the standard -180 to 180 degrees system. With the `t()` function you can rotate a raster object 90 degrees.

Point-to-raster conversion is often done to analyze point-pattern data. For example to count the number of distinct species (represented by point observations) that occur in each raster cell. The `rasterize()` function takes a raster object to set the spatial extent and resolution, and a function to determine how to summarize the points (or an attribute of each point) by cell.

Polygon to raster conversion is typically done to create a `RasterLayer` that can act as a mask, i.e. to set to `NA` a set of cells of a raster object, or to summarize values on a raster by zone. For example a country polygon is transferred to a raster that is then used to set all the cells outside that country to `NA`; whereas polygons representing administrative regions such as states can be transferred to a raster to summarize raster values by region.

It is also possible to convert the values of a raster layer to points or polygons, using `rasterToPoints()` and `rasterToPolygons()`. These functions return values only for cells that are not missing (`NA`).

### Example: Tornado counts by raster

Create a raster of counts indicating the location of all tornado reports since 2003 in 1/2 degree latitude/longitude grid.
```{r}
Torn.sf <- st_read(dsn = "1950-2018-torn-initpoint",
                   layer = "1950-2018-torn-initpoint") %>%
  filter(yr >= 2003)
st_crs(Torn.sf)
```

Next set the raster domain slightly larger than the bounding box and assign a resolution of one degree in longitude and one degree in latitude. Check the extent of the raster with the `extent()` function.
```{r}
frame <- raster(xmn = -106, xmx = -67, 
                ymn = 24, ymx = 50)
res(frame) <- .5
extent(frame)
```

Use the `rasterize()` function to count the number of times each raster cell contains a tornado genesis location. The first argument is the spatial data frame and the second is the raster without values. The argument `field =` specifies a column name in the spatial data frame (here just an identifier) and the argument `fun =` specifies what to do (here simply count the unique instances of the field in each cell). Raster cells without tornadoes are given a value of 0 based on the `background =` argument.

```{r}
Torn.r <- rasterize(x = Torn.sf, 
                    y = frame, 
                    field = "om", 
                    fun = "count",
                    background = 0)
class(Torn.r)
dim(Torn.r)
```

The result is a raster layer. The number of tornadoes occurring in each cell are the values.

```{r}
values(Torn.r)[1:200]
```

To visualize the raster we use the `plot()` method.
```{r}
plot(Torn.r)
```

It looks right. Some cells across the Plains and the South have quite a few tornadoes others not as many.

A spatial statistic indicating how similar values in neighboring cells tend to be is Moran I. It is implemented on rasters with the `Moran()` function. 

```{r}
Moran(Torn.r)
```

Moran I is a global measure of clustering (high values near high values and low values near low values). Values range from -1 to +1 where positive values indicate clustering and negative values indicate regularity (e.g., chessboard).

The estimate of .84 indicates a high level of tornado clustering at this scale. Under the null hypothesis of no spatial autocorrelation the expected value for Moran I is close to zero [-1/(n-1), where n is the number of cells].

With raster data the neighborhood definition is constant across the study region.

Clusters at a local level can be found using a local indicator of spatial autocorrelation. One such indicator is local Moran I, which is computed at each cell (using the `MoranLocal()` function) so the result is a raster.
```{r}
Torn_lmi.r <- MoranLocal(Torn.r)
plot(Torn_lmi.r)
```

Here we can more clearly delineate the hot spots of tornadoes over the south-central Plains.

See also: 
https://rpubs.com/jelsner/tornadoRisk_longTermView
https://rpubs.com/jelsner/tornadoRisk_shortTermView

Focal (neighborhood) functions

The functions: `focal()`, `focalFilter()`, `focalNA()` compute statistics in a neighborhood of cells around a focal cell, putting the result in the focal cell of an output raster. 

With `focal()`, the neighborhood can only be a rectangle. With `focalFilter()`, the neighborhood is a user-defined a matrix of weights and could approximate any shape by giving some cells zero weight. The `focalNA()` function only computes new values for cells that are `NA` in the input raster.

The `distance()` function computes the shortest distance to cells that are not NA. The `pointDistance()` function computes the shortest distance to any point in a set of points. 

The `gridDistance()` function computes the distance when following grid cells that can be traversed (e.g. excluding water bodies). The `direction()` function computes the direction towards (or from) the nearest cell that is not `NA`. The `adjacency()` function determines which cells are adjacent to other cells, and the `pointDistance()` function computes distance between points. 

Summary functions

The function `cellStats()` computes summary statistics on a raster. For example:
```{r}
cellStats(Torn.r, mean)
cellStats(Torn.r, max)
freq(Torn.r)
```

Use `zonal()` to summarize a raster object using zones (areas with the same integer number) defined in a `RasterLayer` and `crosstab()` to cross-tabulate two `RasterLayer` objects.

Convert `raster` to S4 polygons. 
```{r}
TornCounts.sp <- rasterToPolygons(Torn.r)
```

Convert the `SpatialPolygonsDataFrame` to a simple features data frame.
```{r}
TornCounts.sf <- st_as_sf(TornCounts.sp)

ggplot(data = TornCounts.sf) +
  geom_sf(aes(fill = layer))
```