---
title: "Lesson 9"
author: "James B. Elsner"
date: "February 8, 2021"
output:
  html_document: null
editor_options:
  chunk_output_type: console
---

**"Maps invest information with meaning by translating it into visual form."** – Susan Schulten

Today: (1) space time data and (2) making maps 

## Working with space-time data using the {stars} package

In the last lesson we saw how to work with raster data with functions from the {raster} package. Functions from that package require data objects to be of S4 reference class, like those vector data objects in the {sp} package. S4 reference classes allow rich data representations at the expense of flexibility. 

Most packages on CRAN use S3 reference classes because they are more flexible, easier to maintain, and allow for new dialects (e.g., {dplyr}, {ggplot2}). The {stars} package has functions that work with raster data as S3 objects.

Consider a multi-band raster taken from a Landsat 7 view of a small part of the Brazilian coast. It is included in the {stars} package and stored as a _GeoTIFF_ file labeled `L7_ETMs.tif`. We first import it as a raster stack.
```{r}
if(!require(stars)) install.packages("stars", repos = "http://cran.us.r-project.org")
library(stars)

f <- system.file("tif/L7_ETMs.tif",
                  package = "stars")

library(raster)
L7.rs <- stack(f)

class(L7.rs)
```

The data `L7.rs` is a `RasterStack` object as a S4 reference class. 

We list the slot names and extract the extent and CRS using the `@` syntax.
```{r}
L7.rs@extent
L7.rs@crs
```

We extract a single band (layer) from the stack with the `layer =` argument in the `raster()` function. We then plot the raster values with the `plot()` method and compute the spatial autocorrelation with the `Moran()` function as we did in Assignment 4.
```{r}
L7.rB3 <- raster(L7.rs, layer = 3)
plot(L7.rB3)
Moran(L7.rB3)
```

We can convert the raster to a data frame with the `as.data.frame()` method. Here we do that and then compute the normalized difference vegetation index (NDVI) using columns `L7_ETMs.4` and `L7_ETMs.3`.
```{r}
library(tidyverse)

L7.df <- as.data.frame(L7.rs) %>%
  mutate(NDVI = (L7_ETMs.4 - L7_ETMs.3)/(L7_ETMs.4 + L7_ETMs.3))
```

Space-time data often come in the form of arrays, with space and time being array dimensions. Examples include

* raster images
* socio-economic or demographic data,
* environmental variables monitored at fixed stations,
* time series of satellite images with multiple spectral bands,
* spatial simulations, and
* climate and weather model output

The package {stars} provides functions and methods for working with space-time data. To see what methods for class `stars` are available we use the `methods()` function.
``` {r}
methods(class = "stars")
```

The list includes {base} R and {tidyverse} methods.

The canonical data array is that where two dimensions represent spatial raster dimensions and the third a band (or time). [Data array](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube1.png)

But data cubes can have more dimensions. For example, time, space, spectral band, and sensor type. [Data cube](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube2.png)

We import the raster stack as a {stars} object using the `read_stars()` function.
```{r}
L7.stars <- read_stars(f)
L7.stars

dim(L7.stars)
```

In this example, there are three dimensions, two spatial (`x` and `y`), and the third across six bands (`band`). Values across the six bands and space are summarized as a single attribute with name `L7_ETMs.tif`.

The data `L7.stars` is a `stars` object with a S3 reference class. It has no slots and thus no slot names.
```{r}
slotNames(L7.stars)
```

The data are stored in a four dimensional array. The first index is the attribute, the second and third indexes are the spatial coordinates, and the fourth index is the band.

We plot only bands 3 and 4 by subsetting on the fourth index and using the `plot()` method. 
```{r}
plot(L7.stars[,,,3:4])
```

Since this the data object is S3 we can use functions from the `ggplot2()` package together with the `geom_stars()` layer from the {stars} package to plot all 6 bands with a common color scale bar.
```{r}
library(ggplot2)

ggplot() +
  geom_stars(data = L7.stars) +
  facet_wrap(~ band)
```

We can generate a new {stars} object by applying a function to the band values. For example here we again compute NDVI through a function applied across the `x` and `y` spatial dimensions using the `st_apply()` method after creating the function `NVDI()`.
```{r}
NDVI <- function(z) (z[4] - z[3]) / (z[4] + z[3])

( NDVI.stars <- st_apply(L7.stars, 
                         MARGIN = c("x", "y"), 
                         FUN = NDVI) )
ggplot() +
  geom_stars(data = NDVI.stars) 
```

The stars data frame can also be split, here on the band dimension, to yield a representation as six rasters in the list form.
```{r}
( L7split.stars <- split(L7.stars, 
                         f = "band") )
```

Now the bands are given as columns in the data frame part of the {stars} object and there are only two dimensions (`x` and `y`).

Monthly precipitation across the globe

Read a space-time raster stored as a NetCDF file. The data are monthly global precipitation anomalies on 2.5 by 2.5 degree lat/lon grid. We read the NetCDF file using three array dimensions, two planar space, and the third is time (monthly starting in 1948).
```{r}
if(!"precip.mon.anom.nc" %in% list.files()) {
  download.file(url = "http://myweb.fsu.edu/jelsner/temp/data/precip.mon.anom.nc",
                destfile = "precip.mon.anom.nc")
}

( w.stars <- read_stars("precip.mon.anom.nc") )
```

There are two spatial dimensions and the third dimension is time in months.
```{r}
plot(w.stars[,,,1])
```

Raster data do not need to be regular or aligned along the cardinal directions. Functions in the {stars} package supports rotated, sheared, rectilinear and curvi-linear grids. [Grids](https://raw.githubusercontent.com/r-spatial/stars/master/man/figures/README-plot2-1.png)

Functions in the {stars} package also support the vector data model. Vector data cubes arise when we have a single dimension that points to distinct spatial feature geometries, such as polygons (e.g. denoting administrative regions). [Vector data cube polygons](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube3.png)

Or points (e.g., denoting sensor locations). [Vector data cube points](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube4.png)

For more see: https://github.com/r-spatial/stars/tree/master/vignettes and https://awesomeopensource.com/project/r-spatial/stars

Also you can check out some rough code that I've been working on to take advantage of the {stars} functionality including plotting daily temperatures across the U.S. and creating a vector data cube of COVID19 data in the `stars.Rmd` file on course GitHub site.

## Mapping with functions from the {ggplot2} package

As we have already seen, the {ggplot2} package supports `sf` objects with the function `geom_sf()`. An initial `ggplot()` function followed by one or more layers, that are added with `+`. The layers begin with `geom_`.

For example, consider the objects `nz` and `nz_height` from the {spData} package, where `nz` is a simple feature data frame from the New Zealand census with information about the area, population, and sex ratio (male/female) of the 16 districts in the country.
```{r}
library(spData)

str(nz)
```

The simple feature column is labeled `geom` and the geometry type is multi-polygon.

And `nz_height` is a simple feature data frame containing the elevation of specific locations in New Zealand.
```{r}
str(nz_height)
```

The simple feature column is labeled `geomtry` and the geometry type is point.

Here we make a choropleth map of the median income by region across New Zealand and add a layer indicating elevation heights.
```{r}
ggplot() + 
  geom_sf(data = nz, aes(fill = Median_income)) +
  geom_sf(data = nz_height) +
  scale_x_continuous(breaks = c(170, 175))
```

The function `geom_sf()` uses the `geometry` column of the simple feature data frame. The function does not work with S4 spatial data frames.

The function automatically plots graticules (lines of latitude and longitude) labels. The default settings for the graticules can be overridden using `scale_x_continuous()`, `scale_y_continuous()` or `coord_sf(datum = NA)`.

A big advantage of using {ggplot2} for mapping is the large user-community and the many add-on packages. We've already seen {ggrepel} for nice labeling.

The maps from {ggplot2} are given a level of interactivity when printed using the function `ggplotly()` from the {plotly} package. 
```{r}
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
```

For example, let's recreate the choropleth map but this time assigning the object to `g1`. Then to make the map interactive we use `plotly::ggplotly(g1)`.
```{r}
g1 <- ggplot() + 
  geom_sf(data = nz, aes(fill = Median_income)) +
  geom_sf(data = nz_height) +
  scale_x_continuous(breaks = c(170, 175))

plotly::ggplotly(g1)
```

Another example: the county area by state in the U.S. The data as a simple feature data frames is available in the {USAboundariesData} package at `ropensci.org`.
```{r}
if(!require(USAboundariesData)) install.packages("USAboundariesData", repos = "http://packages.ropensci.org", type = "source")

FLcounties.sf <- us_counties(states = "FL")

ggplot(data = FLcounties.sf) +
  geom_sf(aes(fill = aland))
```

But {ggplot2} has a few drawbacks for making maps. The `geom_sf()` function does not always produce a nice legend. Raster objects are not supported and need to be converted into a data frame before plotting.

## Mapping with functions from the {tmap} package

There are other packages for making nice maps with some of them listed in the syllabus. In particular I like the {tmap} package because it is agnostic to the type of spatial data object. S3 simple feature data frames and S4 spatial vector data and raster objects can be used as layers on a single map.
```{r}
if(!require(tmap)) install.packages(pkgs = "tmap", repos = "http://cran.us.r-project.org")
library(tmap)
```

Like functions in the {ggplot2} package, functions in the {tmap} use the 'grammar of graphics'. The grammar separates the input data frame from the aesthetics (how data are made visible). The functions translate the data into aesthetics. The aesthetics can include the location on a geographic map (defined by the geometry), color, and other visual components.

A {tmap} map starts with the `tm_shape()` function that takes as input a spatial data frame. The function is followed by one or more layers such as `tm_fill()`, `tm_dots()`, `tm_raster()`, etc that defines how a property in the data gets translated to an aesthetic. 

Consider again the New Zealand simple feature data frame (`nz`) from the {spData} package.To make a map we first identify the spatial data frame with the `tm_shape()` function and then add a borders layer with the `tm_borders()` layer.
```{r}
tm_shape(nz) +
  tm_borders() 
```

The borders separate New Zealand into 16 administrative regions.

The function `tm_shape()` and its subsequent drawing layers (here `tm_borders()`) as a 'group'. The data in the `tm_shape()` function must be a spatial object of class simple feature, raster, or an S4 class spatial object.

Here we use a fill layer (`tm_fill()`) instead of the borders layer.
```{r}
tm_shape(nz) +
  tm_fill() 
```

The multi-polygons are filled using the same gray color as the borders.

Here we layer using the fill aesthetic and then add a border aesthetic.
```{r}
tm_shape(nz) +
  tm_fill(col = 'green') +
  tm_borders() 
```

Layers are added with the `+` operator and are functionally equivalent to an overlay.

We can assign the resulting map to an object. For example here we assign the map of New Zealand to the object `map_nz`.
```{r}
map_nz <- tm_shape(nz) + 
  tm_polygons()

class(map_nz)
```

The resulting object is of class `tmap`.

New spatial data are added with `+ tm_shape(new_object)`. In this case `new_obj` represents a new spatial data frame to be plotted over the preceding layers. When a new spatial data frame is added in this way, all subsequent aesthetic functions refer to it, until another spatial data frame is added.

For example, let's add an elevation layer to the New Zealand map. The elevation raster (`nz_elev`) spatial data frame is in the {spDataLarge} package on GitHub.

The `install_github()` function from the {devtools} package is used to install packages on GitHub. GitHub is a company that provides hosting for software development version control using Git. Git is a version-control system for tracking changes in code during software development.
```{r}
if(!require(devtools)) install.packages(pkgs = "devtools", repos = "http://cran.us.r-project.org")
library(devtools)

if(!require(spDataLarge)) install_github(repo = "Nowosad/spDataLarge")
library(spDataLarge)
```

Next identify the spatial data for the the new layer by adding `tm_shape(nz_elev)`. Then add the raster layer with the `tm_raster()` function and set the transparency level to 70% (`alpha = .7`).
```{r}
map_nz1 <- map_nz +
  tm_shape(nz_elev) + 
    tm_raster(alpha = .7)

map_nz1
```

The new map object `map_nz1` builds on the existing map object `map_nz` by adding the raster layer `nz_elev` representing elevation.

We can create new layers with functions. For instance, a function like `st_union()` operates on the `geometry` column of a simple feature data frame.

As an example, here we create a line string layer as a simple feature object using three geo-computation functions. We start by creating a union over all polygons (regions) with the `st_union()` function applied to the `nz` simple feature object. The result is a multi-polygon defining the coastlines. 

Then we buffer this multi-polgyon out to a distance of 22.2 km using the `st_buffer()` function. The result is a single polygon defining the coastal boundary around the entire country. 

Finally we change the polygon geometry to a line string geometry with the `st_cast()` function.

The operations are linked together with the pipe operator from the {dplyr} package.
```{r}
nz_water.sfc <- st_union(nz) %>% 
  st_buffer(22200) %>% 
  st_cast(to = "LINESTRING")
class(nz_water.sfc)
```

Now we add the resulting `sfc` as a layer to our map.
```{r}
map_nz2 <- map_nz1 +
  tm_shape(nz_water.sfc) + 
    tm_lines()

map_nz2
```

Finally, lets create a layer representing the country elevation high points (stored in the object `nz_height`) onto the `map_nz2` object with `tm_dots()` function.
```{r}
map_nz3 <- map_nz2 +
  tm_shape(nz_height) + 
    tm_dots()

map_nz3
```

Map layout, facets, and inserts

Layout functions help create a cartographic map. Elements include the title, the scale bar, margins, aspect ratios, etc. For example, here elements such as a north arrow and a scale bar are added with `tm_compass()` and `tm_scale_bar()`, respectively and the `tm_layout()` function is used to add the title and background color.
```{r}
map_nz + 
  tm_compass(type = "8star", 
             position = c("left", "top")) +
  tm_scale_bar(breaks = c(0, 100, 200), 
               text.size = 1) +
  tm_layout(title = "New Zealand",
            bg.color = "lightblue")
```

Faceted maps (referred to as 'small multiples') are composed of several maps arranged side-by-side. Facets enable the visualization of how spatial relationships change with respect to another variable.

For example, here the faceted variable is time (`year`). The simple feature data frame is from the {spData} package. We first filter the data frame keeping only the years 1970, 1990, 2010, and 2030.
```{r}
urb_1970_2030 <- urban_agglomerations %>% 
  filter(year %in% c(1970, 1990, 2010, 2030))
```

Note: The operator `%in%` acts like a recursive or. If year == 1970 or year == 1990, ... For example, 
```{r}
1969:2031 

1969:2031 %in% c(1970, 1990, 2010, 2030)
```

We then make a map.
```{r, eval=FALSE}
tm_shape(world) + 
  tm_polygons() + 
tm_shape(urb_1970_2030) + 
  tm_symbols(col = "black", 
             border.col = "white",
             size = "population_millions") +
  tm_facets(by = "year", 
            nrow = 4, 
            free.coords = FALSE)
```

The above code chunk demonstrates key features of faceted maps created with functions from the {tmap} package.

* Shapes that do not have a facet variable are repeated (the countries in world in this case).
* The `by =` argument which varies depending on a variable (`year` in this case).
* nrow/ncol setting specifying the number of rows (and columns) that facets should be arranged into.
* The `free.coords =` argument specifies whether each map has its own bounding box.

Small multiples are also generated by assigning more than one value to one of the aesthetic arguments. For example here we map the happiness index (`HPI`) and gross domestic product per capita (`gdp_cap_est`).
```{r}
tm_shape(World) +
    tm_polygons(c("HPI", "gdp_cap_est"), 
        style = c("pretty", "kmeans"),
        palette = list("RdYlGn", "Purples"),
        title = c("Happy Planet Index", "GDP per capita")) 
```

Two maps are created each with a different scale. All arguments of the layer functions can be vectorized, one for each small multiple map. Arguments that normally take a vector, such as `palette =`, are placed in a `list()`.

Multiple map objects can also be arranged in a single plot with the `tmap_arrange()` function.
```{r, eval=FALSE}
map1 <- tm_shape(World) +
           tm_polygons("HPI", 
                       style = "pretty",
                       palette = "RdYlGn",
                       title = "Happy Planet Index") 

map2 <- tm_shape(World) +
           tm_polygons("gdp_cap_est", 
                       style = "kmeans",
                       palette = "Purples",
                       title = "GDP per capita") 

tmap_arrange(map1, map2)
```

## Example: Current COVID19 vaccinations by state

Get the data.
```{r}
f <- "https://raw.githubusercontent.com/owid/covid-19-data/e2da3a49250481a8a22f993ee5c3731111ba6958/scripts/scripts/vaccinations/us_states/input/cdc_data_2021-02-06.csv"

df <- read_csv(f) 
```

Get a US census map with geography shifted for Alaska and Hawaii. Rename the state name column (`NAME`) to `LongName`.
```{r}
library(tidycensus)

sf <- get_acs(geography = "state",
              variables = "B01003_001",
              geometry = TRUE,
              shift_geo = TRUE) %>%
  rename(LongName = NAME)
```

Join the COVID data frame with the simple feature data frame from the census. Then make a map showing the doses administered per 100K people as of February 6, 2021.
```{r}
sf <- sf %>%
  left_join(df)

library(tmap)

tm_shape(sf) +
  tm_fill(col = "Admin_Per_100K", title = "Per 100K" ) +
  tm_borders(col = "gray70") +
  tm_layout(legend.outside = TRUE)
```

## Creating an interactive map

A nice feature of {tmap} is that we can create an interactive map using the same code as we used to create a static map. 

For example our static map of New Zealand (`map_nz`) is viewed interactively by switching to view mode.
```{r}
tmap_mode("view")
map_nz
```

With the interactive mode turned on, all maps produced with {tmap} will launch as zoomable HTML. This feature includes the ability to specify the basemap with `tm_basemap()` (or `tmap_options()`) as demonstrated here.
```{r}
map_nz + 
  tm_basemap(server = "OpenTopoMap")
```

We can also create interactive maps with the `tmap_leaflet()` function.

The view mode in {tmap} works with faceted plots. The argument sync in `tm_facets()` is used to produce multiple maps with synchronized zoom and pan settings.
```{r}
world_coffee <- left_join(world, 
                          coffee_data, 
                          by = "name_long")
tm_shape(world_coffee) + 
  tm_polygons(c("coffee_production_2016", 
                "coffee_production_2017")) + 
  tm_facets(nrow = 1, sync = TRUE)
```

Change the view mode back to plot.
```{r}
tmap_mode("plot")
```

## Adding an inset map

An inset map contextualizes the geographic study area. Here we create a map of the central part of New Zealand's Southern Alps. The inset map shows where the main map is in relation to all of New Zealand. 

The first step is to define the area of interest. Here it is done here by creating a new spatial object `nz_region` using the `st_bbox()` function and the `st_as_sfc()` to make it a simple feature column.
```{r}
nz_region <- st_bbox(c(xmin = 1340000, xmax = 1450000,
                       ymin = 5130000, ymax = 5210000),
                     crs = st_crs(nz_height)) %>% 
  st_as_sfc()
```

The second step is to create a base map showing New Zealand's Southern Alps area. This is the closeup view of where the most important message is stated. The region is clipped to the simple feature column `nz_region` created above. The layers include a raster of elevations and locations of high points. A scale bar is included.
```{r}
nz_height_map <- tm_shape(nz_elev, 
                          bbox = nz_region) +
  tm_raster(style = "cont", 
            palette = "YlGn", 
            legend.show = TRUE) +
  tm_shape(nz_height) + 
  tm_symbols(shape = 2, 
             col = "red", 
             size = 1) +
  tm_scale_bar(position = c("left", "bottom"))

nz_height_map
```

The third step is to create the inset map. It gives a context and helps to locate the area of interest. This map clearly indicates the location of the main map.
```{r}
nz_map <- tm_shape(nz) + 
  tm_polygons() +
  tm_shape(nz_height) + 
  tm_symbols(shape = 2, 
             col = "red", 
             size = .1) + 
  tm_shape(nz_region) + 
  tm_borders(lwd = 3)

nz_map
```

The final step is to combine the two maps. The `viewport()` function from the {grid} package is used to give a center location (x and y) and the size (width and height) of the inset map.
```{r}
library(grid)

nz_height_map
print(nz_map, 
      vp = viewport(.8, .27, width = .5, height = .5))
```

Additional details are available here: https://geocompr.robinlovelace.net/adv-map.html