---
title: "Lesson 10"
author: "James B. Elsner"
date: "February 10, 2021"
output:
  html_document: null
editor_options:
  chunk_output_type: console
---

**"You haven't mastered a tool until you understand when it should not be used."** – Kelsey Hightower

Today: Defining neighborhoods and spatial weights. The spatial lag variable and its relationship to autocorrelation.

From last lesson a nice feature of the {tmap} package is that we can create an interactive map using the same code as we used to create a static map. 

For example create map showing the county borders in Florida.
```{r}
library(tmap)
library(USAboundaries)
library(sf)

FLcounties.sf <- us_counties(state = "FL",
                             resolution = "high")

( FLmap <- tm_shape(FLcounties.sf) +
  tm_borders() )
```

Toggle the mode to `view` with the `tmap_mode()` function.
```{r, eval=FALSE}
tmap_mode("view")
FLmap
tmap_mode("plot")
```

## Defining neighborhoods and neighborhood weights

The concept of autocorrelation plays a central role in spatial statistics. It measures the degree to which things tend to cluster. Things include attribute values aggregate to polygons (or raster cells) or spatial locations. How it gets estimated depends on the geometry of the spatial data.

Clustering arises from:

* Association: whatever is causing an attribute to have a certain value in one area causes the attribute to have a similar value in areas nearby. Crime rates in nearby neighborhoods might tend to cluster due to similar factors such as economic status and the amount of policing. Non-infectious diseases (e.g., lung cancer) might have similar rates in neighborhoods close to an oil refinery.

* Causality: something within a given area directly influences outcomes within nearby areas. The broken window theory of crime suggests that poverty, lack of maintenance, and petty crime tends to breed more crime due to a perceived breakdown in civil order.

* Interaction: the movement of people, goods or information creates relationships between areas. Infectious diseases might spread from a source region thus increasing the disease rates in surrounding areas as the direct result of contact or movement of people between regions.

Spatial statistics quantify, and condition on, autocorrelation but they are silent about physical causes. Understanding the reason for autocorrelation is important for inference because the causal mechanism might be confounded by its. Divorce rates are high in states in the South, but so is the number of Waffle Houses. Understanding causation requires domain specific knowledge.

When a variable's values are aggregated to regions, autocorrelation is quantified by calculating how similar a value in region $i$ is to the value in region $j$ and weighting this similarity by how 'close' region $i$ is to region $j$. Closer regions have greater weight.

High similarities with high weight (similar values close together) yield high values of spatial autocorrelation. Low similarities with high weight (dissimilar values close together) yield low values of spatial autocorrelation. Let $\hbox{sim}_{ij}$ denote the similarity between values $Y_i$ and $Y_j$, and let $w_{ij}$ denote a weight describing the 'distance' between regions $i$ and $j$, for $i$, $j$ = 1, ..., $N$. 

Then a spatial autocorrelation index (SAI) is given by
$$
\hbox{SAI} = \frac{\sum_{i,j=1}^N w_{ij}\hbox{sim}_{ij}}{\sum_{i,j=1}^N w_{ij}}
$$
which represents the weighted similarity between regions. The collection of weights is called a spatial weights matrix. The spatial weights matrix defines the neighbors for each region and their strength of association.

For cells in a raster under the rook-contiguity criterion, $w_{ij}$ = 1 if cell $i$ and $j$ share a boundary, and 0 if they don't share a boundary. In this case $w_{ij}$ = $w_{ji}$. Also, a cell is not a neighbor of itself so $w_{ii}$ = 0.

Alternatively we can define center locations from a set of polygon regions and let $w_{ij}$ = 1 if the center of region $i$ is near the center of region $j$ and 0 otherwise. Here we need to decide on the number of nearest neighbors.

We can also define neighbors by distance. For example, if $d_{ij}$ is the distance between centers $i$ and $j$, we can let $w_{ij}$ = 1 if $d_{ij}$ < $\delta$ and 0 otherwise.

Consider crime data at the tract level in the city of Columbus, Ohio (Anselin, 1988: Spatial Econometrics. Boston, Kluwer Academic). The tract polygons are projected with arbitrary spatial coordinates.
```{r}
if(!"columbus" %in% list.files()) {
download.file(url = "http://myweb.fsu.edu/jelsner/temp/data/columbus.zip",
              destfile = "columbus.zip")
unzip("columbus.zip")
}

( CC.sf <- read_sf(dsn = "columbus") )
```

The simple feature data frame contains housing values (`HOVAL`), income values (`INC`) and (`CRIME`) in census tracts across the city of Columbus, OH.

Create a choropleth map of the crime rates (`CRIME`). Residential burglaries and vehicle thefts per 1000 households.
```{r}
library(tmap)

tm_shape(CC.sf) +
  tm_fill(col = "CRIME",
          title = "") +
  tm_borders(col = "gray70") +
  tm_layout(title = "Burglary & Vehicle Thefts\n/1000 Households",
            legend.outside = TRUE)
```

Note that the variable name must be in quotes.

Alternatively
```{r, eval=FALSE}
library(ggplot2)

ggplot(data = CC.sf) + 
  geom_sf(aes(fill = CRIME), col = "gray70") +
  ggtitle("Burglary & Vehicle Thefts/1000 Households") +
  theme_void()
```

High crime areas tend to cluster. Autocorrelation quantifies this clustering. To quantify we first need to define a list of neighbors for each polygon.

We create a list of neighbors using the `poly2nb()` function from the {spdep} package. The function now works on S4 spatial classes and simple features.
```{r}
if(!require(spdep)) install.packages("spdep", repos = "http://cran.us.r-project.org")
library(spdep)

( nbs <- poly2nb(CC.sf) )
```

The `nb` in the `poly2nb()` function names stands for neighbor list object. The function builds the list from geometries based on contiguity. Neighbors must share one or more points. By default the contiguity is defined as having at least one point in common. This is changed by using the argument `queen = FALSE`.

There are 49 tracts (polygons). Each tract is bordered by at least one other tract. The average number of neighbors is 4.8. The total number of neighbors over all tracts is 236. This represents 9.8% of all possible connections (if every tract is a neighbor of itself and every other tract 49 * 49).

A graph of the neighbor links is obtained with the `plot()` method. The arguments include the neighbor list object (`nbs`) and the location of the polygon centers, which are extracted from the simple feature data frame using the `st_centroid()`.
```{r}
plot(CC.sf$geometry)
plot(nbs, 
     st_centroid(CC.sf$geometry),
     add = TRUE)
```

The graph is a network (topology) showing the contiguity pattern (adjacency neighbor structure). Tracts close to the center of the city have the most neighbors.

The number of links per tract (node)--link distribution--is obtained with the `summary()` method.
```{r}
summary(nbs)
``` 

The list of neighboring tracts for the first two tracts.
```{r}
nbs[[1]]
nbs[[2]]
```

The first tract has two neighbors that include tracts 2 and 3. The neighbor numbers are stored as an integer vector. Tract 2 has three neighbors that include tracts 1, 3, and 4. Tract 5 has 8 neighbors and so on. The function `card()` tallies the number of neighbors by tract.
```{r}
card(nbs)
```

Tract 5 has 8 neighbors.

Next, weights are added to the neighbor list object to create a weights matrix. The weights specifying how 'close' each neighbor is. This is done with the `nb2listw()` function. The function turns the neighbor list object into a spatial weights object. By default the weighting scheme gives each link the same weight equal to the multiplicative inverse of the number of neighbors.
```{r}
wts <- nb2listw(nbs)

class(wts)
```

This new `wts` object is a list with two elements. The first element (`listw`) is the weights matrix and the second element (`nb`) is the neighbor list object.
```{r}
summary(wts)
```

The network statistics are given along with information about the weights. By default all neighboring tracts are assigned a weight equal to the inverse of the number of neighbors  (`style = "W"`). For a tract with 5 neighbors each neighbor gets a weight of 1/5. The sum over all weights (`S0`) is the number of tracts.

To see the weights for the first two tracts type
```{r}
wts$weights[1:2]
```

The `$weights` list a sparse representation of the spatial weights matrix. The full matrix has dimensions 49 x 49.

To see the neighbors of the first two tracts type
```{r}
wts$neighbours[1:2]
```

Tract 1 has two neighbors (tract 2 & 3) so each are given a weight of 1/2. Tract 2 has three neighbors (tract 1, 3, & 4) so each are given a weight of 1/3.

With the weights matrix specified and saved as an object we are ready to compute a metric of spatial autocorrelation.

Caution: Contiguity can result in areas having no neighbors; islands for example. By default the `nb2listw()` function assumes each area has at least one neighbor. If this is not the case we need to specify how areas without neighbors are handled using the argument `zero.policy = TRUE`. This permits the weights list to be formed with zero-length weights vectors.

For example, consider the districts in Scotland.
```{r}
download.file("http://myweb.fsu.edu/jelsner/temp/data/scotlip.zip",
              "scotlip.zip")
unzip("scotlip.zip")

SL.sf <- read_sf(dsn = "scotlip", 
                 layer = "scotlip")
plot(SL.sf$geometry)
```

Here there are three island districts.

Create a list of neighbors.
```{r}
( nbs2 <- poly2nb(SL.sf) )
```

Three regions with no links.

Use the `nb2listw()` function with the argument `zero.policy = TRUE`. Otherwise we get an error saying the empty neighbor sets are found.
```{r}
wts2 <- nb2listw(nbs2,
                 zero.policy = TRUE)
head(wts2$weights)
```

## Computing autocorrelation

A commonly used autocorrelation statistic is Moran I. Moran I follows the basic form of spatial autocorrelation indexes where the similarity between regions $i$ and $j$ is proportional to the product of the deviations from the mean such that
$$
\hbox{sim}_{ij} \propto (Y_i - \bar Y) (Y_j - \bar Y)
$$

where $i$ indexes the region and $j$ indexes the neighbors of $i$. The value of $\hbox{sim}_{ij}$ is large when the $Y$ values in the product are on the same side of their respective means and small when they are on opposite sides of their respective means.

The formula for I is
$$
\hbox{I} = \frac{N} {W} \frac {\sum_{i,j} w_{ij}(Y_i-\bar Y) (Y_j-\bar Y)} {\sum_{i} (Y_i-\bar Y)^2}
$$
where $N$ is the number regions, $w_{ij}$ is the matrix of spatial weights, and $W$ is the sum over all weights.

Consider the following spatial grid of cells containing random attribute values.
```{r}
if(!require(spatstat)) install.packages(pkgs = "spatstat", repos = "http://cran.us.r-project.org")
library(spatstat)

set.seed(6750)
Y <- ppp(runif(200, 0, 1), 
         runif(200, 0, 1))
plot(quadratcount(Y), main = "")
```

The formula for I results in one value for the entire grid. 

Let's first consider a single cell on the grid ($N$ = 1). Here the middle cell (row 3, column 3). Let $i$ = 3 in the above formula and let $j$ index the cells touching the center cell in reading order starting with cell (2, 2), then cell (2, 3), etc.

Assume each neighbor is given a weight of 1/8 so $W$ = 1. Then the value of I for the single center cell is 
I_{3, 3} = (6 - mean(y)) * 
                     ((8 - mean(y)) + 
                     (3 - mean(y)) +
                     (9 - mean(y)) +
                     (12 - mean(y)) +
                     (10 - mean(y)) +
                     (10 - mean(y)) +
                     (9 - mean(y))) / 
            (6 - mean(y))^2)

```{r}
y <- c(3, 10, 7, 12, 5, 11, 8, 3, 9, 12, 
      6, 12, 6, 10, 3, 8, 10, 10, 9, 7, 
      5, 10, 8, 5, 11)
( yb <- mean(y) )
```

```{r}
Inum_i <- (6 - yb) * 
                 ((8 - yb) + (3 - yb) + (9 - yb) + 
                  (12 - yb) + (10 - yb) + (10 - yb) + 
                  (10 - yb) + (9 - yb))
Iden_i <- (6 - yb)^2
Inum_i/Iden_i
```

The I value of -3.5 indicates that the center cell which has a value below the average over all 25 cells is mostly surrounded by cells having values above the average.

Repeat this calculation for every cell and then take the sum.

This is what the function `moran()` from the {spdep} package does. The first argument is the vector containing the values for which we are interested in determining the magnitude of the spatial autocorrelation and the second argument is the `listw` object. 

Further, we need to specify the number of regions and the sum of the weights `S0`. The latter is obtained from the `Szero()` function applied to the `listw` object.

Returning to the Columbus crime data here we let `m` be the number of census tracts and `s` be the sum of the weights. We then apply the `moran()` function on the variable `CRIME` in
```{r}
m <- length(CC.sf$CRIME)
s <- Szero(wts)

moran(CC.sf$CRIME, 
      listw = wts, 
      n = m, 
      S0 = s)
```

The function returns the Moran I statistic and the kurtosis (K) of the distribution of crime values. Moran I ranges from -1 to +1. The value of .5 for the crime rates indicates fairly high spatial autocorrelation. This is expected based on the clustering of crime in the central city. Positive values of Moran I indicate clustering and negative values indicate inhibition.

Kurtosis is a statistic measuring the peakedness of the distribution of the attribute values. A normal distribution has a kurtosis of 3. If the kurtosis is too large (or small) relative to a normal distribution then the inferences we make with Moran I are suspect.

Another index of spatial autocorrelation is the Geary C statistic. The equation is 
$$
\hbox{C} = \frac{(N-1) \sum_{i,j} w_{ij} (Y_i-Y_j)^2}{2 W \sum_{i}(Y_i-\bar Y)^2} 
$$
where again $W$ is the sum over all weights ($w_{ij}$) and $N$ is the number of areas.

The syntax of the `geary()` function is similar to the syntax of the `moran()` function except we also specify `n1` to be one minus the number of polygons.
```{r}
geary(CC.sf$CRIME, 
      listw = wts,
      n = m, 
      S0 = s, 
      n1 = m - 1)
```

Values for Geary C range from 0 to 2 with 1 indicating no spatial autocorrelation. Values less than 1 indicate positive autocorrelation. Both I and C are global measures of autocorrelation, but C is more sensitive to local variations in autocorrelation.

Rule of thumb: If the interpretation of Geary C is much different than the interpretation of Moran I then consider computing local measures of spatial autocorrelation.

## The spatial lag variable and its relation to autocorrelation

Moran I is the slope coefficient from a regression of the weighted average of the neighborhood values onto the observed values. The weighted average of neighborhood values is called the spatial lag.

Let `crime` be the set of crime values in each region. We create a spatial lag variable using the `lag.listw()` function. The first argument is the `listw` object and the second is the vector of crime values.
```{r}
crime <- CC.sf$CRIME
Wcrime <- lag.listw(wts, 
                    crime)
```

For each value in the vector `crime` there is a corresponding value in the vector `Wcrime` representing the average crime over the neighboring regions.

Recall tract 1 had tract 2 and 3 as its only neighbors. So the following should return a `TRUE`.
```{r}
Wcrime[1] == (crime[2] + crime[3])/2
```

A scatter plot of the neighborhood average crime rates versus the individual polygon crime rates in each shows there is a relationship.
```{r}
library(dplyr)
library(ggplot2)

data.frame(crime, Wcrime) %>%
ggplot(aes(x = crime, y = Wcrime)) + 
  geom_point() + 
  geom_smooth(method = lm) +
  scale_x_continuous(limits = c(0, 70)) +
  scale_y_continuous(limits = c(0, 70)) +
  xlab("Crime") + 
  ylab("Average Crime in the Neighborhood") +
  theme_minimal()
```

The vertical axis contains the neighborhood average crime rate. The range of neighborhood averages is smaller than the range of individual polygon crime rates.

Tracts with low values of crime tend to be surrounded by tracts with low values of crime on average and tracts with high values of crime tend be surrounded by tracts with high values of crime. The slope is upward (positive).

The magnitude of the slope is the Moran I value. To check this type
```{r}
lm(Wcrime ~ crime)
```

The coefficient on the `crime` variable in the simple linear regression is .5. 

The scatter plot is a  Moran scatter plot.

Let's consider another data set.
```{r}
download.file("http://myweb.fsu.edu/jelsner/temp/data/sids2.zip",
              "sids2.zip")
unzip("sids2.zip")

SIDS.sf <- read_sf(dsn = "sids2")
st_crs(SIDS.sf) <- 4326
head(SIDS.sf)
```

The column `SIDR79` contains the death rate per 1000 live births (1979-84) from sudden infant death syndrome. Create a choropleth map of the SIDS rates.
```{r}
tm_shape(SIDS.sf) +
  tm_fill("SIDR79", title = "") +
  tm_borders(col = "gray70") +
  tm_layout(title = "SIDS Rates 1979-84 [per 1000]",
            legend.outside = TRUE)
```

Again, alternatively we can use `geom_sf()`.
```{r, eval=FALSE}
ggplot(data = SIDS.sf) + 
  geom_sf(aes(fill = SIDR79), col = "gray70") +
  ggtitle("SIDS Rates 1979-84 [per 1000]") +
  theme_void()
```

Create a neighborhood list (`nb`) and a `listw` object (`wts`) then graph the neighborhood network.
```{r}
nbs <- poly2nb(SIDS.sf)
wts <- nb2listw(nbs)

plot(nbs, 
     st_centroid(st_geometry(SIDS.sf)))
```

Next compute Moran I on the SIDS rates over the period 1974--1979.
```{r}
m <- length(SIDS.sf$SIDR79)
s <- Szero(wts)

moran(SIDS.sf$SIDR79, 
      listw = wts, 
      n = m, 
      S0 = s)
```

I is .14 and K is 4.4. A normal distribution has a kurtosis of 3. Values less than about 2 or greater than about 4 indicate that inferences about autocorrelation based on the assumption of normality are suspect.

Weights are specified using the `style =` argument in the `nb2listw()` function. The default "W" is row standardized (sum of the weights over all links equals the number of polygons). "B" is binary (each neighbor gets a weight of one). "S" is a variance stabilizing scheme. 

Each style gives a somewhat different value for I.
```{r}
x <- SIDS.sf$SIDR79
moran.test(x, nb2listw(nb, style = "W"))$estimate[1]
moran.test(x, nb2listw(nb, style = "B"))$estimate[1]  # binary
moran.test(x, nb2listw(nb, style = "S"))$estimate[1]  # variance-stabilizing
```

When we report a Moran I we need to state what type of weighting was used.

Let `sids` be a vector with elements containing the SIDS rate in each county. We create a spatial lag variable using the `lag.listw()` function. The first argument is the `listw` object and the second is the vector of rates.
```{r}
sids <- SIDS.sf$SIDR79
Wsids <- lag.listw(wts, 
                   sids)
```

For each value in the vector `sids` there is a corresponding value in the object `Wsids` representing the neighborhood average SIDS rate. 
```{r}
Wsids[1]
j <- wts$neighbours[[1]]
j
sum(SIDS.sf$SIDR79[j])/length(j)
```

The weight for county one is `Wsids[1]` = 2.659. The neighbor indexes for this county are in the vector `wts$neighbours[[1]]` of length 3. Add the SIDS rates from those counties and divide by the number of counties (`length(j)`).

A scatter plot of the neighborhood average SIDS rate versus the actual SIDS rate in each region.
```{r}
data.frame(sids, Wsids) %>%
ggplot(aes(x = sids, y = Wsids)) + 
  geom_point() + 
  geom_smooth(method = lm) +
  scale_x_continuous(limits = c(0, 7)) +
  scale_y_continuous(limits = c(0, 7)) +
  xlab("SIDS") + ylab("Spatial Lag of SIDS") +
  theme_minimal()
```

The regression line slopes upward indicating positive spatial autocorrelation. The value of the slope is I. To check this type
```{r}
lm(Wsids ~ sids)
```