---
title: "Lesson 13"
author: "James B. Elsner"
date: "February 22, 2021"
output:
  html_document: null
editor_options:
  chunk_output_type: console
---

Let's see how to choose between these two options with an example. Returning to the Columbus crime data, we import the data, fit a linear regression model to statistically explain crime rates using income and housing values. Further we first need to check whether a spatial regression model is warranted by testing whether there is significant spatial autocorrelation in the residuals of the aspatial model.
```{r}
if(!"columbus" %in% list.files()){
download.file("http://myweb.fsu.edu/jelsner/temp/data/columbus.zip",
              "columbus.zip")
unzip("columbus.zip")
}

CC.sf <- read_sf(dsn = "columbus", 
                 layer = "columbus")
model.lm <- lm(CRIME ~ INC + HOVAL, 
               data = CC.sf)
CC.sf$residuals <- residuals(model.lm)

nbs <- poly2nb(CC.sf)
wts <- nb2listw(nbs)

lm.morantest(model.lm, wts)
```

The answer is 'yes' a spatial regression model is warranted.

To help decide between the two spatial regression models described above we run a sequence of statistical tests on our linear model object. The tests are the Lagrange multiplier (LM) tests. LM refers to a technique to determine the coefficients on the explanatory variables while simultaneously determining the coefficient on the spatial regression term. It does this iteratively.

The tests are performed with the `lm.LMtests()` function. The test type is specified as a character string. The tests should be considered in order. Start with the standard version of both the spatial error model and spatial lag model LM tests.

For example, to perform a LM test for the spatial error and spatial lag model on the Columbus crime model we type
```{r}
lm.LMtests(model.lm, 
           listw = wts, 
           test = c("LMerr", "LMlag"))
```

The output shows that both the spatial error and spatial lag model terms are significant ($p$-value < .15). Ideally one term is significant and the other is not and we choose the model with the significant term.

Since both are significant we should test again. This time we use the robust forms of the statistics.
```{r}
lm.LMtests(model.lm, 
           listw = wts, 
           test = c("RLMerr", "RLMlag"))
```

Here the error model has a large $p$-value and the lag model has a $p$-value that is less than .15 so we choose the lag model for our spatial regression.

A decision tree (from Luc Anselin) shows the sequence of tests for making a choice about which type of spatial model to use [Decision Tree](http://myweb.fsu.edu/jelsner/temp/SpatialRegressionDecisionTree.png)

If both tests show significance for both models, then we can fit both models and check which one results in the lowest AIC.

### Model fitting

To fit a spatial lag model to the crime data we use the `lagsarlm()` function. The value for $\rho$ is found first using the `optimize()` function and then the $\beta$'s are obtained using generalized least squares. These functions are now in the {spatialreg} package.
```{r}
if(!require(spatialreg)) install.packages(pkgs = "spatialreg", repos = "http://cran.us.r-project.org")

model.lag <- spatialreg::lagsarlm(CRIME ~ INC + HOVAL, 
                      data = CC.sf, 
                      listw = wts)

summary(model.lag)
```