---
title: "Lesson 17"
author: "James B. Elsner"
date: "March 8, 2021"
output:
  html_document: null
editor_options:
  chunk_output_type: console
---

**"To me programming is more than an important practical art. It is also a gigantic undertaking in the foundations of knowledge."** – Grace Hopper

## Intensity trend as a possible confounding factor

Quantifying the report bias works well for Kansas where there is no trend in the local intensity. There is no large differences in the number of reports as we move across different regions of the state.

Things are different in Texas. Convert the tornado reports (EF1 or worse) occurring over the state as a `ppp` object. We use an EPSG code or a Texas-centric Lambert conic conformal projection.
```{r}
Torn.sf <- st_read(dsn = "1950-2018-torn-initpoint") %>%
  st_transform(crs = 3082) %>%
  filter(mag >= 0)

T.ppp <- Torn.sf %>%
  as_Spatial() %>%
  as.ppp()

W <- us_states(states = "Texas") %>%
  st_transform(crs = st_crs(Torn.sf)) %>%
  as_Spatial() %>%
  as.owin()

( T.ppp <- T.ppp[W] %>%
  spatstat::rescale(s = 1000, 
                    unitname = "km") )
intensity(T.ppp)
```

There are 8,736 tornado reports. The distance unit is kilometer. The average intensity is .013 events per square kilometer over this 69-year period (1950-2018).

Next we plot the local intensity using a kernel smoother.
```{r}
T.ppp %>%
  density() %>%
  plot()
```

There is a clear trend of tornado reports from a low in the southwest to a high in the northeast. The average intensity of .013 tornadoes per square km is too high in southwestern parts of the state and too low in the northern parts.

Next we compute and plot the spatial intensity as a smoothed function of distance to nearest town or city. We start by removing the marks on the tornado events assigning the unmarked `ppp` object to `Tum.ppp`. We then create a `ppp` object from the city/town locations and subset the tornado events by the window.
```{r}
Tum.ppp <- T.ppp %>%
  unmark()

C.ppp <- C.sf %>%
  st_transform(crs = st_crs(Torn.sf)) %>%
  as_Spatial() %>%
  as.ppp() %>%
  unmark()

C.ppp <- C.ppp[W] %>%
  spatstat::rescale(s = 1000,
                    unitname = "km")
```

Next create a distance map of the city/town locations using the `distmap()` function.
```{r}
Zc <- distmap(C.ppp)
plot(Zc)
```

Finally we compute the intensity of tornadoes as a smoothed function of distance to nearest town/city with the `rhohat()` function. We then prepare the output and make a plot.
```{r}
rhat <- rhohat(Tum.ppp, 
               covariate = Zc,
               method = "transform")

data.frame(dist = rhat$Zc, 
           rho = rhat$rho, 
           hi = rhat$hi, 
           lo = rhat$lo) %>%
ggplot() +
  geom_ribbon(aes(x = dist, ymin = lo , ymax = hi), alpha = .3) +
  geom_line(aes(x = dist, y = rho), color = "red") +  
  scale_y_continuous(limits = c(0, NA)) +
  geom_hline(yintercept = intensity(Tum.ppp), color = "blue") +
  ylab("Tornado reports per sq. km") +
  xlab("Distance from nearest town center (km)") +
  theme_minimal()
```

We see that the intensity of the tornado reports is much higher than the average intensity in the vicinity of towns and cities. However caution needs to exercised in the interpretation because the trend of increasing tornado reports moving from southwest to northeast across the state mirrors the trend in the occurrence of cities/towns.

We can quantify this effect by specifying a function in the `covariate =` argument. Here we specify a flat surface with `x,y` as arguments and `x + y` inside the function.
```{r}
plot(rhohat(Tum.ppp, 
            covariate = function(x,y){x + y},
            method = "transform"))
```

Local intensity increases along the axis labeled `X` starting at a value of 7,400. At value of `X` equal to about 8,200 the spatial intensity stops increasing.

Units along the horizontal axis are kilometers but the reference (intercept) distance is at the far left. So we interpret the increase in spatial intensity going from southwest to northeast as a change across about 800 km (8200 - 7400).

The local intensity of cities has the same property (increasing from southwest to northeast then leveling off).
```{r}
plot(rhohat(C.ppp, 
            covariate = function(x,y){x + y},
            method = "transform"))
```

So the population bias towards more reports near towns/cities is potentially confounded by the fact that there tends to be more cities and towns in areas that have conditions more favorable for tornadoes.

Thus we can only get so far by examining intensity estimates if our interest lies in inferring the cause of spatial variation in the intensity. We will need to look at second order properties of the events.

## Given a Texas tornado, what is the chance that it will cause at least EF3 damage?

Combining maps allow us to map estimates of relative risks of events. More generally the relative risk is a conditional probability. For example, given a tornado in Texas what is the chance that it will cause at least EF3 damage? 
```{r}
library(sf)
library(tidyverse)
library(spatstat)
library(maptools)
library(USAboundaries)

Torn.sf <- st_read(dsn = "1950-2018-torn-initpoint") %>%
  st_transform(crs = 3082) %>%
  filter(mag >= 0) %>%
  mutate(EF = as.factor(mag)) %>%
  dplyr::select(EF)

T.ppp <- Torn.sf %>%
  as_Spatial() %>%
  as.ppp()

W <- us_states(states = "Texas") %>%
  st_transform(crs = st_crs(Torn.sf)) %>%
  as_Spatial() %>%
  as.owin()

T.ppp <- T.ppp[W] %>%
  spatstat::rescale(s = 1000, 
                    unitname = "km")
summary(T.ppp)
```

There are 8,736 Texas tornadoes of which 368 are EF3+ on the damage-rating scale. The distance unit is kilometer. The average intensity is .013 events per square kilometer over this 69-year period (1950-2018).

For the state as a whole we have the answer from our summary of the `ppp` object.
```{r}
summary(T.ppp)
```

The chance that a tornado anywhere in Texas will be at least EF3 or worse is the sum of the proportions for these types: .03594 + .00549 + .00069 = .042 (or 4.2\%). Or the sum of the intensities for these types divided by the overall intensity (1.264744e-08). 

But as we saw the intensity varies spatially.

We create two `ppp` objects the first one being the set of all tornado locations with damage ratings 0, 1, or 2 and the other the set of all tornado locations with damage ratings 3, 4, or 5.

First we split the object then merge them and assign names as marks.
```{r}
H.ppp <- unmark(T.ppp[T.ppp$marks == 2 | T.ppp$marks == 1 | T.ppp$marks == 0])
I.ppp <- unmark(T.ppp[T.ppp$marks == 3 | T.ppp$marks == 4 | T.ppp$marks == 5])
T2.ppp <- superimpose(H = H.ppp, I = I.ppp)
```

The probability that a tornado chosen at random is intense (EF3+) is 4%. Plot touchdown locations for the set of intense tornadoes.
```{r}
plot(I.ppp, pch = 25, cols = "red", main="")
plot(T.ppp, add=TRUE, lwd = .1)
```

To obtain the relative risk we use the `relrisk()` function. If X is a bivariate point pattern (a multitype point pattern consisting of two types of events) then by default, the events of the first type (the first level of `marks(X)`) are treated as controls or non-events, and events of the second type are treated as cases. 

Then the function `relrisk()` computes the spatially-varying probability of a case, (i.e. the probability $p(u)$ that a point at location $u$ will be a case). If `relative = TRUE`, it computes the spatially-varying relative risk of a case relative to a control, $r(u) = p(u)/(1 - p(u))$.

Here we compute the relative risk on a 128 by 128 grid. It takes a few seconds.
```{r}
rr <- relrisk(T2.ppp, 
              dimyx = c(128, 128))
```

The result is again an object of class `im` (a pixel object with values we can interpret as the conditional probability of an 'intense' tornado, see https://en.wikipedia.org/wiki/Enhanced_Fujita_scale).

We retrieve the range of probabilities with the `range()` function. Note that many of the values are `NA` corresponding pixels that are outside the window so we set `na.rm = TRUE`.
```{r}
range(rr, na.rm = TRUE)
```

The probabilities range from a low of .77% to a high of 6.1%.

County borders from the {map} package (automatically installed with {ggplot2}). It provides maps of the USA, with state and county borders, that can be retrieved and converted as sf objects.
```{r}

TX.sf <- us_counties(state = "TX") %>%
  st_transform(crs = st_crs(Torn.sf))
``` 

Create a map. To facilitate plotting the results we convert the resulting `im` object to a raster and set the CRS accordingly.
```{r}
library(raster)

rr.r <- raster(rr)
crs(rr.r) <- st_crs(Torn.sf)$proj4string

library(tmap)

tm_shape(rr.r) +
  tm_raster(title = "Probability") +
tm_shape(TX.sf) +
  tm_borders(col = "gray70") +
tm_layout(frame = FALSE) +
  tm_credits(text = "Chance that a tornado\ndoes at least EF3 damage",
             size = 1,
             position = c("left", "bottom")) 
```

It is of considerable interest to extract these probabilities for specific cities.

Here we use the data frame `us.cities` from the {map} package has a list of US cities with population greater than about 40,000. Also included are state capitals of any population size.
```{r}
Cities.sf <- us_cities(state = "TX") %>%
  st_transform(crs = st_crs(Torn.sf))
```

We use the `extract()` function from the {raster} package to get a single value for each city. We put these values into the simple feature data frame. 
```{r}
Cities.sf$rr <- raster::extract(rr.r, 
                                Cities.sf)

Cities.sf %>%
  dplyr::arrange(desc(rr)) 
```

To put the finishing touch on this analysis we create a chart using the `geom_lollipop()` function from the {ggalt} package.
```{r}
library(ggalt)
library(scales)

Cities.sf <- Cities.sf %>%
  dplyr::filter(rr > .042)

ggplot(Cities.sf, aes(x = reorder(name, rr), y = rr)) +
    geom_lollipop(point.colour = "steelblue", point.size = 3) +
    scale_y_continuous(labels = percent) +
    coord_flip() +
    labs(x = "", y = NULL, 
         title = "Chance that a random tornado will do at least EF3 damage",
         subtitle = "Cities in Texas with a 2010 population > 40,000",
         caption = "Data from SPC") +
  theme_minimal()
```

## Given a wildfire in Florida what is the probability is was started by lightning?

The spatial wildfire occurrence data for the United States, 1992-2015 [FPA_FOD_20170508]. Data publication contains GIS data (Karen C. Short). It is available here: https://www.fs.usda.gov/rds/archive/catalog/RDS-2013-0009.4. 

I download the GPKG data. GeoPackage is an open, standards-based, platform-independent, portable, self-describing, compact format for transferring geospatial information. Of the 8 million or so wildfires in the U.S. over the period 1992-2015 I extracted those occurring in Florida and saved it as a ESRI Shapefile.

We import the Florida wildfire data as a simple feature data frame and transform the native CRS to a Florida GDL Albers (EPSG 3086).
```{r}
if(!"FL_Fires" %in% list.files()){
  download.file("http://myweb.fsu.edu/jelsner/temp/data/FL_Fires.zip",
                "FL_Fires.zip")
unzip("FL_Fires.zip")
}

FL_Fires.sf <- st_read(dsn = "FL_Fires") %>%
  st_transform(crs = 3086)
dim(FL_Fires.sf)
```

There are over 90K events in Florida. To make things run faster here we analyze a random sample of the events. We first create a random sample of 2000 row numbers and assign the row numbers to the vector `index`. We then subset the `FL_Fires.sf` sf data frame by this index.
```{r}
set.seed(78732)

FL_FiresS.sf <- FL_Fires.sf %>%
  sample_n(size = 2000)

dim(FL_FiresS.sf)
```

The variable `STAT_CAU_1` indicates the cause.
```{r}
table(FL_FiresS.sf$STAT_CAU_1)
```

We then convert the simple feature event set to a `ppp` object over a window defined by the state boundaries with the cause as a factor mark.
```{r}
F.ppp <- FL_FiresS.sf["STAT_CAU_1"] %>%
  as_Spatial() %>%
  as.ppp()

W <- us_states(states = "Florida") %>%
  st_transform(crs = st_crs(FL_Fires.sf)) %>%
  as_Spatial() %>%
  as.owin()

F.ppp <- F.ppp[W]

marks(F.ppp) <- as.factor(marks(F.ppp)) # make the character marks factor marks

summary(F.ppp)
```

The probability that a wildfire is caused by lightning is about 25% (`proportion` column of the frequency versus type table). How does this probability vary over the state?

First we split the object and then merge them and assign names as marks.
```{r}
L.ppp <- F.ppp[F.ppp$marks == "Lightning"] %>%
  unmark()

NL.ppp <- F.ppp[F.ppp$marks != "Lightning"] %>%
  unmark()

LNL.ppp <- superimpose(NL = NL.ppp, 
                       L = L.ppp)
```

The function `relrisk()` computes the spatially-varying probability of a case, (i.e. the probability $p(u)$ that a point at location $u$ will be a case).

Here we compute the relative risk on a 256 by 256 grid.
```{r}
wfr <- relrisk(LNL.ppp, 
               dimyx = c(256, 256))
```

We map the raster as before first converting the image object to a raster object and assigning the CRS. We add the county borders for geographic reference.
```{r}
wfr.r <- raster(wfr)

crs(wfr.r) <- st_crs(FL_Fires.sf)$proj4string

FL.sf <- us_counties(state = "FL") %>%
  st_transform(crs = st_crs(FL_Fires.sf))

library(tmap)

tm_shape(wfr.r) +
  tm_raster(title = "Probability") +
tm_shape(FL.sf) +
  tm_borders(col = "gray70") +
tm_legend(position = c("left", "center") ) +
tm_layout(main.title = "Chance a wildfire was started by lightning (1992-2015)",
          main.title.size = 1) +
tm_compass(position = c("right", "top")) +
tm_credits(text = "Data source: Karen Short https://doi.org/10.2737/RDS-2013-0009.4",
           position = c("left", "bottom")) 
```