---
title: "Lesson 25"
author: "James B. Elsner"
date: "April 9, 2021"
output:
  html_document: null
editor_options:
  chunk_output_type: console
---

**"Statistics is such a powerful language for describing data in ways that reveal nothing about their causes. Of course statistics is powerful for revealing causes as well. But it takes some care. Like the difference between talking and making sense."** -- Richard McElreath

## Interpolation using functions from the {gstat} package

We start with the Meuse river data from Burrough and McDonnell (1998). The data are part of {gstat}.

Convert to {sf} object first? See Tropical cyclone rainfall example.
```{r}
library(gstat)
library(sp)

data(meuse)
head(meuse)
```

The data are topsoil heavy metal concentrations along with a number of soil and landscape variables at locations in a flood plain of the river Meuse near the town of Stein (NL).

Heavy metal concentrations are composite samples of an area of approximately 15 m x 15 m. The dataset also gives local elevation (m) above local river bed and distance to river normalized to a value between 0 and 1.

Create a spatial points data frame (S4 class) and map the concentration of zinc in the soil on a log scale as a bubble plot.
```{r}
coordinates(meuse) <- ~ x + y
proj4string(meuse) <- CRS("+init=epsg:28992")
class(meuse)
bubble(meuse, "zinc", 
       do.log = TRUE, 
       key.space = "bottom")
```

Zinc in the soil is concentrated along the east bank of the Meuse river.

### Inverse-distance weighting

We begin by creating an interpolated surface with the method of inverse distance weighting (IDW). Here the average of the observed values for any location is weighted inversely based on how far the observation points are from the location of interest. An exponent on the weights matrix determines the degree to which nearer observations are weighted more than more distant ones. 

A regular grid where the interpolated values of zinc are made is given in the data frame `meuse.grid`. 
```{r}
data(meuse.grid)
head(meuse.grid)
```

Convert the grid as a data frame to a spatial pixels data frame. Note this is done first by converting the data frame to spatial points then to spatial pixels.
```{r}
coordinates(meuse.grid) <- ~ x + y
proj4string(meuse.grid) <- CRS("+init=epsg:28992")
meuse.grid2 <- as(meuse.grid, "SpatialPixelsDataFrame") 
head(meuse.grid2)
```

As mentioned last time, the above conversion works since the locations have a spatial ordering. It will not work for a data frame with an arbitrary ordering of locations. 

Here we plot the pixel locations using the attribute `dist` indicating distance to the river.
```{r}
library(tmap)

tm_shape(meuse.grid2) +
  tm_raster("dist")
```

Interpolate the zinc values to the pixel locations using IDW. The `idp = ` argument is the exponent on the weights. The larger the exponent, the more variation there is in the interpolated values.
```{r}
zinc.idw <- idw(zinc ~ 1, 
                meuse, 
                newdata = meuse.grid2, 
                idp = 2.5)
str(zinc.idw, max.level = 3)
```

Note the formula. Here `zinc ~ 1`, where `zinc` is the field variable that is to be interpolated and 1 indicates a IDW interpolation (no autocorrelation and no trend or covariates).

The saved object (`zinc.idw`) inherits the class of the interpolation grid (here `SpatialPixelDataFrame`). The interpolated values are saved in the data frame slot with variable name `var1.pred`. The other variable is `var1.var` which is an estimate of the interpolation errors. With IDW they are `NA`.

IDW results in a surface that is often similar to a kriged surface. But interpolations may show undesired effects if the observation locations have a non-uniform distribution.

Map the results. Start with an image plot, then add a bubble plot of the observed concentrations and a legend. Here the size of the point character (circle) is set by the `cex = argument`.
```{r}
tm_shape(zinc.idw) +
  tm_raster("var1.pred") +
tm_shape(meuse) +
  tm_bubbles("zinc", alpha = .3)
```

IDW tends to generate 'bulls-eye' patterns.

The spatial class makes it easier to create maps and export them to different platforms.

### Create a geoTIF file that can be read into ENVI or ArcMap.

Here you repeat the analysis but this time use a planar projection for the interpolation grid, interpolate the logarithm of the zinc, and create a `geoTIF` file for output.

Again make `meuse.grid` a spatial pixels data frame.
```{r}
data(meuse.grid)
coordinates(meuse.grid) <- ~ x + y
gridded(meuse.grid) <- TRUE
proj4string(meuse.grid) <- CRS(paste("+init=epsg:28992", 
                                     "+towgs84=565.237,50.0087,465.658,-0.406857,0.350733,-1.87035,4.0812"))
```

Make `meuse` a spatial points data frame with the same projection.
```{r}
data(meuse)
coordinates(meuse) <- ~ x + y
proj4string(meuse) <- CRS(proj4string(meuse.grid))
```

Use the `krige` function in the {gstat} package and perform inverse-distance predictions. Since no variogram is specified in the function, the method defaults to inverse-distance weighting.
```{r}
logZinc <- krige(log(zinc) ~ 1, 
                meuse, 
                meuse.grid)["var1.pred"]
summary(logZinc)
```

### Create a KML file that can be read by Google Earth [Extra]

We export a colored raster of interpolated log zinc ppm values to a PNG file with an alpha channel for viewing it in Google Earth. Since GE requires geographical coordinates, a number of steps are needed.

First we make a polygon to bound the study area and project it to geographical coordinates.
```{r}
library(maptools)
grd <- as(meuse.grid, "SpatialPolygons")
proj4string(grd) <- CRS(proj4string(meuse.grid))
grd.union <- unionSpatialPolygons(grd, rep("x", length(slot(grd, "polygons"))))  
ll <- CRS("+proj=longlat +datum=WGS84")
grd.union.ll <- spTransform(grd.union, ll)
plot(grd.union.ll, axes = TRUE) 
```

The result is a single polygon as the union of all polygons in grd in geographic coordinates. Next, we construct a grid in geographical coordinates as our target object for export using the `GE_SpatialGrid` function. This grid is the container for the output PNG graphics file. `GE_SpatialGrid` returns values that will be used in setting up the PNG graphics device.

We use the `over()` method to set grid cells outside the river bank area to NA and then discard them by coercion.
```{r}
llGRD <- GE_SpatialGrid(grd.union.ll)
llGRD_in <- over(llGRD$SG, grd.union.ll)
llSGDF <- SpatialGridDataFrame(grid = slot(llGRD$SG, "grid"), 
                              proj4string = CRS(proj4string(llGRD$SG)), 
                              data = data.frame(in0 = llGRD_in))
llSPix <- as(llSGDF, "SpatialPixelsDataFrame")
```

Next we interpolate. Here again we use the `idw` function to make an inverse-distance weighted interpolation of zinc values in ppm.
```{r}
meuse_ll <- spTransform(meuse, 
                       CRS("+proj=longlat +datum=WGS84"))
llSPix$pred <- idw(log(zinc) ~ 1, 
                   meuse_ll, 
                   llSPix)$var1.pred
spplot(llSPix["pred"])
```

Since `GE_SpatialGrid` was set to the size of an PNG graphics device, we can use it with the `image()` function. Other base graphics methods and functions can be used to create an image overlay. First open the graphics device, then set the plotting margins, then use `image`, and finally turn off the device.
```{r}
png(file <- "zincIDW.png", width = llGRD$width, 
    height = llGRD$height, bg = "transparent")
par(mar = c(0, 0, 0, 0), xaxs = "i", yaxs = "i")
image(llSPix, "pred", col = bpy.colors(20))
dev.off()
```

After closing the graphics device, we use the `kmlOverlay` to write the interpolated grid to a KML file giving the location of the overlay. This will load the image at the position when opened in Google Earth.
```{r}
kmlOverlay(llGRD, "zincIDW.kml", "zincIDW.png")
```

Open the `zincIDW.kml` file in Google Earth.

### Distance to river as a covariate

Here you use distance to river as a covariate. It is a trend model for the spatial variability of zinc using a covariate defined by the square root of the distance to river. This is not kriging since no variogram model is used. 
```{r}
zn.lm <- krige(log(zinc) ~ sqrt(dist), 
               locations = meuse, 
               newdata = meuse.grid)

tm_shape(zn.lm) +
  tm_raster("var1.pred")
```

Note that the spacing in colors is not linear.

### Empirical variogram analysis

The `variogram()` function from {gstat} computes the empirical variogram. Here the trend is not in the cardinal directions but rather in distance to the river. The variogram is on the residuals from this trend.
```{r}
meuse.v <- variogram(log(zinc) ~ sqrt(dist), 
                     data = meuse)
plot(meuse.v, xlab = "Lag Distance (m)")
```

The default cutoff lag (`cutoff`) and lag tolerance (`width`) can be changed. For example, type
```{r}
plot(variogram(log(zinc) ~ sqrt(dist), 
               data = meuse, 
               cutoff = 1000, 
               width = 50), 
      xlab = "Lag Distance (m)")
```

For comparison, plot a variogram of residuals after removing a 1st-order trend in direction.
```{r}
plot(variogram(log(zinc) ~ x + y, 
               data = meuse, 
               cutoff = 1000, 
               width = 50), 
     xlab = "Lag Distance (m)")
```

Directional variograms are estimated by adding the argument alpha.
```{r}
plot(variogram(log(zinc) ~ x + y, 
               data = meuse, 
               cutoff = 1000, 
               width = 50, 
               alpha = c(0, 45, 90, 135)), 
     xlab = "Lag Distance (m)")
```

The variogram in the sw-ne direction appears to be different. And there appears to be very little spatial autocorrelation in the n-s direction. This is partly due to the anistropy of the domain.

### Variogram maps

Another way of looking at directional dependence in semivariograms is with variogram maps. Instead of classifying point pairs $Z(s)$ and $Z(s + h)$ by direction and distance separately, they are classified jointly. 

If $h = (x, y)$ is the two-dimensional coordinates of the separation vector, in the variogram map the semivariance contribution of each point pair $(Z(s) - Z(s + h))^2$ is attributed to the grid cell in which $h$ lies. The map is centered at (0, 0), as $h$ is geographical distance rather than geographical location. 

Cutoff and width correspond to map extent and cell size; the semivariance map is point symmetric around (0, 0), as $\gamma(h) = \gamma(-h)$. 

The variogram map is obtained with the `map = TRUE` argument. The threshold assures that only semivariogram map values based on at least 5 point pairs are shown, removing too noisy estimation. 
```{r}
meuse.vmap <- variogram(log(zinc) ~ x + y, 
                        data = meuse, 
                        cutoff = 1500, 
                        width = 100,
                        map = TRUE)
plot(meuse.vmap, threshold = 5)
```

The symmetry about the dx, dy = (0, 0) is apparent. The variances range from 0 to 2.5 and appear to elongate along the long axis of the domain.

Compare with a variogram map when the square root of the distance to river is used as a covariate.
```{r}
meuse.vmap <- variogram(log(zinc) ~ sqrt(dist), 
                        data = meuse, 
                        cutoff = 1500, 
                        width = 100,
                        map = TRUE)
plot(meuse.vmap, threshold = 5)
```

The variances range from 0 to .8 and the directional dependence is much less obvious in this case.

### Variogram modeling

A plot of the model families is available with the `show.vgms()` function.
```{r}
show.vgms() 
```

We've seen the exponential (`Exp`), spherical (`Sph`), and Gaussian (`Gau`) families when working with the functions in the {geoR} package.

After choosing a family starting values are needed in the `fit.variogram()` function. The function calls the `vgm()` function that includes initial guesses for the partial sill (`psill`), the range (`range`), and the nugget (`nugget`) and uses a weighted least squares method.
```{r}
meuse.vm <- fit.variogram(meuse.v, 
                          vgm(model = "Sph", 
                              psill = .15,
                              range = 800, 
                              nugget = .1))
meuse.vm
```

The output includes a 2 x 2 table with the first row giving information about the nugget and the second giving information about the rest of the model. Here the nugget is .08. By definition the range is zero. The partial sill is .15 with a range of 872.7 m. Plot the sample variogram and the variogram model.
```{r}
plot(meuse.v, meuse.vm, 
     xlab = "Lag distance (m)")
```

The `fit.variogram` function refines the initial estimates given in the `vgm` function. If our initial choice is poor the fit may not work.  For example try
```{r}
fit.variogram(meuse.v, 
              vgm(model = "Sph",
                  psill = 1, 
                  range = 10, 
                  nugget = 1))
```

An argument for visual fitting over numerical fitting is when you have knowledge that goes beyond the information in the data. Information may come from other studies or derived from measurement error characteristics for a specific measuring device (nugget effect).

### Kriging

Simple kriging is done when there is no trend and the mean height of the surface is provided. Because you know the mean exactly you also know the residuals exactly at the data locations. If you know the residuals exactly then you can do a better job of estimating the autocorrelation function (variogram). For ordinary kriging you estimate the mean so the residuals are also estimates.

Universal kriging is done when there is a trend. With universal kriging both the trend and spatial variability are estimated in tandem.

Here we apply kriging to the logarithm of the zinc concentration using the fitted variogram model.
```{r}
lz.sk <- krige(formula = log(zinc) ~ 1, 
               locations = meuse, 
               newdata = meuse.grid, 
               model = meuse.vm, 
               beta = 5.9)
lz.ok <- krige(formula = log(zinc) ~ 1, 
               locations = meuse, 
               newdata = meuse.grid, 
               model = meuse.vm)
lz.uk <- krige(formula = log(zinc) ~ sqrt(dist),
               locations = meuse, 
               newdata = meuse.grid, 
               model = meuse.vm)

m1 <- tm_shape(lz.sk) +
        tm_raster("var1.pred", title = "Simple")

m2 <- tm_shape(lz.ok) +
        tm_raster("var1.pred", title = "Ordinary")

m3 <- tm_shape(lz.uk) +
        tm_raster("var1.pred", title = "Universal")

tmap_arrange(m1, m2, m3, ncol = 3)
```

Simple and ordinary kriging treat all spatial variation as autocorrelation. Universal kriging lets some of the variation be due to trend and covariates.

Predictive skill.

## Example 1: Tropical cyclone rainfall

In 2008, Tropical Cyclone Fay formed from a tropical wave near the Dominican Republic, passed over the island of Hispaniola, Cuba, and the Florida Keys, then crossed the Florida peninsula and moved westward across portions of the panhandle producing heavy rains in parts of the state.

Storm total rainfall amounts from stations in and around the state are in `FayRain.txt`. They are a compilation of reports from the U.S. NOAA/NWS official weather sites and coop sites.

The coop sites are the Community Collaborative Rain, Hail and Snow Network (CoCoRaHS), a community-based, high density precipitation network made up of volunteers who take measurements of precipitation in their backyards. The data were obtained from NOAA/NCEP/HPC and from the Florida Climate Center.

Load the data.
```{r}
L <- "http://myweb.fsu.edu/jelsner/temp/data/FayRain.txt"
FR.df <- read.table(L, header = TRUE)
names(FR.df)
```

The data frame contains 803 rainfall sites. Longitude and latitude coordinates of the sites are given in the first two columns and total rainfall in inches and millimeters are given in the second two columns. Create a spatial points data frame by specifying columns that contain the spatial coordinates. Then assign a geographic coordinate system and convert the rainfall from millimeters to centimeters.
```{r}
library(sf)

FR.sf <- st_as_sf(x = FR.df,
                 coords = c("lon", "lat"),
                 crs = 4326)

FR.sf$tpm <- FR.sf$tpm/10
summary(FR.sf$tpm)
```

The median value is 15.8 cm and the maximum is 60.2 cm. 

Next get the Florida county boundaries from the {USAboundaries} package.
```{r}
FL.sf <- USAboundaries::us_counties(states = "Florida")
```

Next we create a character string specifying the tags for a planar projection and transform the geographic coordinates of the site locations and map polygons to the projected coordinates. Here we use Florida GDL Albers (EPSG:3087) with meter as the distance unit.

```{r}
FR.sf <- st_transform(FR.sf, crs = 3087)
FL.sf <- st_transform(FL.sf, crs = 3087)
```

A map of the rainfall sites and storm totals with the state boundaries is made by typing
```{r}
library(tmap)

tm_shape(FR.sf) +
  tm_dots(col = "tpm", size = 1) +
tm_shape(FL.sf) +
  tm_borders()
```

Two areas of heavy rainfall are noted.  One running north-south along the east coast and another across the north. Rainfall collection sites are clustered in and around cities. This will make it difficult to use a spline interpolation.

### Empirical variogram

Rainfall is an example of geostatistical data. In principle it can be measured anywhere, but typically we have values at a sample of sites. The pattern of observation sites is not of much interest as it is a consequence of constraints (convenience, opportunity, economics, etc) unrelated to the phenomenon. Interest centers on inference about how much rain fell across the region.

The empirical variogram is computed using the `variogram()` function. The first argument is the model formula specifying the rainfall column from the data frame and the second argument is the data frame name.  Here `~ 1` in the model formula indicates no covariates or trends in the data. Trends can be included by specifying coordinate names through the `st_coordinates()` function.

We compute the empirical variogram for Fay's rainfall and save it by typing
```{r}
FR.v <- variogram(tpm ~ 1, 
                  data = FR.sf)
```

We plot the variogram values as a function of lag distance and add text indicating the number of point pairs for each lag distance.
```{r}
library(ggplot2)

v.df <- data.frame(dist = FR.v$dist/1000,
                   gamma = FR.v$gamma,
                   np = FR.v$np)

( pv <- ggplot(v.df, aes(x = dist, y = gamma)) +
  geom_point() +
  geom_text(aes(label = np), nudge_y = -5) +
  scale_y_continuous(limits = c(0, 220)) +
  scale_x_continuous(limits = c(0, 400)) +
  xlab("Lagged distance (h) [km]") +
  ylab(expression(paste("Semivariance (", gamma, ") [", cm^2, "]"))) +
  theme_minimal() )
```

Values start low (~50 cm$^2$) at short lag distance, then increase to over 200 cm$^2$ at lag distance of about 200 km.

The zero-lag semivariance is called the 'nugget' and the semivariance at a level where the variogram values no longer increase is called the 'sill.'  The lag distance to the sill is called the 'range.'  These three parameters (nugget, sill, and range) are used to fit a model to the variogram.

### Variogram model

Next we fit a model to the empirical variogram. The variogram model is a mathematical relationship defining the semivariance as a function of lag distance. We first save the family and the initial parameter guesses in a variogram model (`FR.vmi`) object by typing
```{r}
FR.vmi <- vgm(model = "Gau", 
              psill = 150, 
              range = 200 * 1000, 
              nugget = 50)
FR.vmi
```

The `psill` argument is the partial sill as the difference between the sill and the nugget. We get estimates of the parameter values from looking at the empirical variogram. 

Next we use the `fit.variogram()` function to improve the fit.  Given a set of initial parameter values, the method of weighted least squares improves the parameter estimates. Ordinary least squares is not appropriate as the semivariances are correlated across the lag distances and the precision on the estimates varies depending on the number of site pairs for a given lag.
```{r}
FR.vm <- fit.variogram(object = FR.v, 
                       model = FR.vmi)
FR.vm
```

The result is a variogram model with a nugget of 46.8 cm$^2$, a partial sill of 157 cm$^2$, and a range on the sill of 129 km. 

Let $r$ be the range, $c$ the partial sill and $c_o$ the nugget, then the equation defining the curve over the set of lag distances $h$ is
$$
\gamma(h)=c\left(1-\exp\left(-\frac{h^2}{r^2}\right)\right)+c_o
$$

We create a data frame with values of h and gamma using this equation.

```{r}
nug <- FR.vm$psill[1]
ps <- FR.vm$psill[2]
r <- FR.vm$range[2] / 1000
h <- seq(0, 400, .2)
gamma <- ps * (1 - exp(-h^2 / (r^2))) + nug

vm.df <- data.frame(dist = h,
                    gamma = gamma)

pv + geom_line(aes(x = dist, y = gamma), data = vm.df)
```

Check for anisotropy.
```{r}
plot(variogram(tpm ~ 1, 
               data = FR.sf, 
               alpha = c(0, 45, 90, 135)), 
     xlab = "Lag Distance (m)")
```

We see the range of correlations is longer (about 300 km) in the north-south direction (0 degrees). We refit the variogram defining an anistropy ellipse with the `anis =` argument. The first parameter is the direction of longest range and the second parameter is the ratio of the longest to shortest. Here about (200/300 = .67).
```{r}
FR.vmi <- vgm(model = "Gau", 
              psill = 150, 
              range = 300 * 1000, 
              nugget = 50,
              anis = c(0, .67))
FR.vm <- fit.variogram(FR.v, FR.vmi)
```

### Kriging

The final step is to use the variogram model together with the rainfall values at the observation sites to create an interpolated surface. The process is called kriging. As Edzer Pebesma notes, 'krige' is to 'kriging' as 'predict' is to 'predicting.' Here we use ordinary kriging as there are no spatial trends in the rainfall.

Interpolation is done using the `krige()` function. The first argument is the model specification and the second is the data. Two other arguments are needed. One is the variogram model using the argument name model and the other is a set of locations identifying where the interpolations are to be made. This is specified with the argument name `newdata =`.

#### Interpolate to point loctions

Here we interpolate to locations on a regular grid. We create a grid of locations within the boundary of Florida using the `st_sample()` function.
```{r}
grid.sf <- st_sample(FL.sf,
                     size = 5000,
                     type = "regular")
```

We specify the number of locations using the argument `size =`. Note that the actual number of locations will be somewhat different because of the irregular boundary. 

First we use the `krige()` function to interpolate the observed rainfall to the grid locations. Recall that for a given location, the interpolation is a weighted average of the rainfall across the entire region where the weights are determined by the variogram model.
```{r}
ipl <- krige(tpm ~ 1, 
             locations = FR.sf, 
             newdata = grid.sf,
             model = FR.vm)
```

If the variogram model is not included then inverse distance weighted interpolation is performed. The function will not work if different values share the same location. 

The saved object (`ipl`) inherits the spatial object specified in the `newdata` argument, but extends it to a spatial data frame. The data frame with two variables.  The first `var1.pred` is the interpolated rainfall and the second `var1.var` is the prediction variance.

We plot the interpolated field.
```{r}
tm_shape(ipl) +
  tm_dots("var1.pred",
          size = .1,
          palette = "Greens",
          title = "Rainfall (cm)") +
  tm_shape(FL.sf) +
  tm_borders() +
  tm_layout(legend.position = c("left", "bottom"),
            title = "Tropical Cyclone Fay (2008)",
            title.position = c("left", "bottom"))
```

Note that a portion of the data values where outside of the state but our interest is only to have values on a grid within the state. 

The spatial interpolation shows that parts of east central and north Florida were deluged by Fay.

#### Interpolate to the counties (block kriging)

We use block kriging to estimate average rainfall within each county. The county-wide rainfall average is relevant for water resource managers. Block kriging produces a smoothed estimate of this area average, which will differ from a simple average over all sites within the county because of spatial autocorrelation.

We use the same function to interpolate but specify the spatial polygons rather than the spatial grid as the new data.
```{r}
ipl2 <- krige(tpm ~ 1, 
              locations = FR.sf, 
              newdata = FL.sf, 
              model = FR.vm)
```

Again we plot the interpolations.
```{r}
tm_shape(ipl2) +
  tm_polygons(col = "var1.pred",
            palette = "Greens",
            title = "Rainfall (cm)") +
  tm_layout(legend.position = c("left", "bottom"),
            title = "Tropical Cyclone Fay (2008)",
            title.position = c("left", "bottom"))
```

The overall pattern of rainfall from Fay featuring the largest amounts along the central east coast and over the eastern panhandle are similar in both maps.

We compare the kriged average with the simple average at the county level with `aggregate()` from the {sf} package.
```{r}
ipl3 <- aggregate(FR.sf, by = FL.sf, FUN = mean)
```

The function returns a simple feature data frame of the average rainfall in each county.

The state-wide mean of the kriged estimates at the county level is 
```{r}
round(mean(ipl2$var1.pred), 2)
```

This compares with a state-wide mean from the simple averages.
```{r}
round(mean(ipl3$tpm), 2)
```

The correlation between the two estimates across the 67 counties is 
```{r}
round(cor(ipl3$tpm, ipl2$var1.pred), 2)
```

The variogram model reduces the standard deviation of the kriged estimate relative to the standard deviation of the simple averages because of the local smoothing.
```{r}
round(sd(ipl2$var1.pred), 2)
round(sd(ipl3$tpm), 2)
```

This can be seen with a scatter plot of simple averages versus kriged averages at the county level.
```{r}
compare.df <- data.frame(simpleAvg = ipl3$tpm,
                         krigeAvg = ipl2$var1.pred)
ggplot(compare.df, aes(x = simpleAvg,
                       y = krigeAvg)) +
  geom_point() +
  geom_abline(slope = 1) +
  geom_smooth(method = lm, se = FALSE)
```

### Uncertainty

Another advantage of kriging as a method of spatial interpolation is the accompanying uncertainty estimates.  

The prediction variances are listed in a column in the spatial data frame saved from apply the `krige()` function. Variances are smaller in regions with more rainfall observations.  

Prediction variances are also smaller with block kriging as much of the variability within the county averages out. To compare the distribution characteristics of the prediction variances for the point and block kriging of the rainfall observations, type
```{r}
round(summary(ipl$var1.var), 1)
round(summary(ipl2$var1.var), 1)
```

The median prediction variance (in cm$^2$) for our point kriging is close to the value of the nugget.
```{r}
round(fivenum(ipl$var1.var)[3], 1)
```

In contrast the median prediction variance for our block kriging is a much smaller 
```{r} 
round(fivenum(ipl2$var1.var)[3], 1)
```

Simulations exploit this uncertainty and provide synthetic data for use in deterministic models. 

Conditional simulation, where the simulated field (realization) is generated given the data and the variogram model, is done using the same `krige()` function by adding the argument `nsim` to specify the number of simulations.  

For a large number it may be necessary to limit the number neighbors in the kriging. This is done using the `nmax` argument. For a given location, the weights assigned to observations far away are very small, so it is efficient to limit how many are used in the simulation.

As an example, here we generate four realizations of the county-level storm total rainfall for Fay and limit the neighborhood to 50 of the closest observation sites. Note that it may take a few seconds.
```{r}
ipl.sim <- krige(tpm ~ 1, 
                 locations = FR.sf, 
                 newdata = FL.sf, 
                 model = FR.vm, 
                 nsim = 4, 
                 nmax = 50)
```

Simulations are conditional on the observed rainfall and the variogram model using block kriging on the counties.

```{r}
library(tmap)

tm_shape(ipl.sim) +
    tm_polygons(col = c("sim1", "sim2", "sim3", "sim4"),
                palette = "Greens",
                title = "Simulated Rainfall [cm]") +
    tm_facets(free.scales = FALSE) 
```

The overall pattern of rainfall remains the same, but there are differences especially in counties with fewer observations and in counties where the rainfall gradients are sharp.


## Example 2: Beauregard Tornado

BeauregardTornado.Rmd

## Cokriging

Statistical interpolation can be extended to obtain surfaces of multiple field variables. The idea is that if two field variables are correlated then information about the spatial correlation in one field variable can help provide information about values in the other field variable. The spatial variability of one variable is correlated with the spatial variability of the other variable. And this idea is not limited to only two variables. 

Here we return to the Meuse River data. Concentrations of heavy metals in the flood plain of the Meuse River in Holland.
```{r}
library(gstat)
library(sp)

data(meuse)
names(meuse)
coordinates(meuse) <- ~ x + y
prj <- paste("+init=epsg:28992", 
             "+towgs84=565.237,50.0087,465.658,-0.406857,0.350733,-1.87035,4.0812")
proj4string(meuse) <- CRS(prj)
```

Suppose we are interested in the spatial distribution of all the heavy metals in the soils along the river (Not just zinc like we looked in Lesson 22). 

First we need to organize the data as a `gstat` object. This is done with the `gstat()` function which orders (and copies) the field variables into a single object. It is done successively with the `gstat()` function.

Here we specify the trend using the square root of the distance to river as we did previously.
```{r}
g <- gstat(NULL, "logCd", log(cadmium) ~ sqrt(dist), meuse)
g <- gstat(g, "logCu", log(copper) ~ sqrt(dist), meuse)
g <- gstat(g, "logPb", log(lead) ~ sqrt(dist), meuse)
g <- gstat(g, "logZn", log(zinc) ~ sqrt(dist), meuse)
g
```

Next we use the `variogram()` function to compute empirical variograms. The function, when operating on a `gstat` object, computes all direct and cross variograms.
```{r}
v <- variogram(g)
plot(v)
```

The plot method displays the set of direct and cross variograms. The direct variograms are shown in the four panels along the diagonal of the triangle of plots.

The cross variograms are shown in the six panels below the diagonal. For example, the cross variogram between the values of cadmium and copper is given in the second row of the first column. 

Q: The cross variogram is similar to what function from the `spatstat` package?

Next we use `fit.lmc()` to fit separate models to each of the empirical variograms. We use an initial partial sill and nugget equal to one and a range of 800.
```{r}
vm <- fit.lmc(v, g, 
              vgm(psill = 1, model = "Sph", 
              range = 800, nugget = 1))
plot(v, vm)
```

As the variograms indicate, the variables have a strong cross correlations. Because these variables are co-located, we can also compute direct correlations.
```{r}
cor(as.data.frame(meuse)[c("cadmium", "copper", "lead", "zinc")])
```

The correlation matrix confirms strong cross correlation among the four variables at zero lag. The cross variogram generalizes these correlations across lag distance.

Given the variogram models, cokriged maps are produced using the `predict()` method after setting the grid locations for interpolation. The CRS for the grid locations matches the CRS of the data.
```{r}
data(meuse.grid)
coordinates(meuse.grid) <- ~ x + y
proj4string(meuse.grid) <- CRS(proj4string(meuse))
cok <- predict(vm, meuse.grid)
names(cok)
```

The predictions for logarithm of zinc concentration are plotted.
```{r}
spplot(cok, "logZn.pred")
```

Compare with predictions using only zinc.
```{r}
v2 <- variogram(log(zinc) ~ sqrt(dist), 
               data = meuse)
vm2 <- fit.variogram(v2, vgm(psill = .15, model = "Sph", 
                             range = 800, nugget = .1))
uk <- krige(log(zinc) ~ sqrt(dist), meuse, newdata = meuse.grid, 
              model = vm2)
spplot(uk, "var1.pred")
```

The predicted covariances between zinc and cadmium are plotted.
```{r}
spplot(cok, "cov.logCd.logZn")
```

The map shows areas of the flood plain with high (and low) correlations between cadmium and zinc.

Obtaining a quality statistical interpolation is a nuanced process but with practice kriging can be an important tool in a spatial analyst's toolbox.

## Assignment

Consider the data set `parana` from the {geoR} package containing rainfall values from the state of Paran\`a. The data are average rainfall (mm) over the period May-June.

1. Plot the values. Plot the residuals after a 1st order trend is removed.
2. Compute and plot an empirical variogram on the residuals after removing the 1st-order trend. Check for anisotropy.
3. Fit a variogram model from the gaussian family to the empirical variogram. Assume isotropy. Set the initial values of the nugget to 500, the partial sill to 800 and the range to 100. Plot the modeled variogram on top of the empirical variogram.
4. Create a kriged surface by interpolating the rainfall values to locations on a regular grid. Use the `pred_grid(parana$borders, by = 2)` function to generate the grid.
5. Plot the interpolated surface. Note this can be done with the `image()` function. Use `col = rev(terrain.colors(21))`. 
6. Compute the cross-validated prediction errors assuming a constant variogram model. Plot the observed versus predicted values as a scatter plot. Include the best fit line and the y = x line.

## Block cross validation for species distribution models

See `BCVforSDM.Rmd`.